Malaria parasites require a divergent heme oxygenase for apicoplast gene expression and biogenesis

Version 1

Abstract

Abstract

Malaria parasites have evolved unusual metabolic adaptations that specialize them for growth within heme-rich human erythrocytes. During blood-stage infection, Plasmodium falciparum parasites internalize and digest abundant host hemoglobin within the digestive vacuole. This massive catabolic process generates copious free heme, most of which is biomineralized into inert hemozoin. Parasites also express a divergent heme oxygenase (HO)-like protein (PfHO) that lacks key active-site residues and has lost canonical HO activity. The cellular role of this unusual protein that underpins its retention by parasites has been unknown. To unravel PfHO function, we first determined a 2.8 Å-resolution X-ray structure that revealed a highly α-helical fold indicative of distant HO homology. Localization studies unveiled PfHO targeting to the apicoplast organelle, where it is imported and undergoes N-terminal processing but retains most of the electropositive transit peptide. We observed that conditional knockdown of PfHO was lethal to parasites, which died from defective apicoplast biogenesis and impaired isoprenoid-precursor synthesis. Complementation and molecular-interaction studies revealed an essential role for the electropositive N-terminus of PfHO, which selectively associates with the apicoplast genome and enzymes involved in nucleic acid metabolism and gene expression. PfHO knockdown resulted in a specific deficiency in levels of apicoplast-encoded RNA but not DNA. These studies reveal an essential function for PfHO in apicoplast maintenance and suggest that Plasmodium repurposed the conserved HO scaffold from its canonical heme-degrading function in the ancestral chloroplast to fulfill a critical adaptive role in organelle gene expression.

Introduction

Malaria remains a devastating infectious disease marked by increasing treatment failures with frontline artemisinin therapies1,2. Plasmodium malaria parasites diverged early in eukaryotic evolution from well-studied yeast and mammalian cells, upon which most understanding of eukaryotic biology is based. Due to this evolutionary divergence, parasites acquired unusual molecular adaptations for specialized growth and survival within human erythrocytes, human hepatocytes, and mosquitos. Unraveling and understanding these adaptations will provide deep insights into the evolution of Plasmodium and other apicomplexan parasites and unveil new parasite-specific vulnerabilities that are suitable for therapeutic targeting to combat increasing parasite drug resistance.

Heme metabolism is central to parasite survival within red blood cells (RBCs), the most heme-rich cell in the human body. During blood-stage growth, Plasmodium parasites internalize and digest up to 80% of host hemoglobin within the acidic digestive vacuole (DV)3,4. This massive digestive process liberates an excess of cytotoxic free heme that is detoxified in situ within the DV via biomineralization into chemically inert hemozoin crystals. Other hematophagous organisms and cells, including certain blood-feeding insects and human liver/splenic macrophages that process senescent RBCs, depend on canonical heme oxygenase (HO) enzymes to degrade excess heme and recycle iron5–7. In contrast to these examples, Plasmodium parasites lack an active HO pathway for enzymatic heme degradation and rely fully on alternative mechanisms for heme detoxification and iron acquisition during blood-stage infection8. Nevertheless, malaria parasites express a divergent HO-like protein (PfHO, Pf3D7_1011900) with unusual biochemical features9. Although this protein shows distant homology to HO enzymes, it lacks the strictly conserved active-site His ligand and does not degrade heme in vitro or in live cells8. Genome-wide knockout and insertional mutagenesis studies in P. berghei and P. falciparum reported that the PfHO gene was refractory to disruption10,11, but the biological function that underpins the retention and putative essentiality of this divergent HO-like protein in malaria parasites has remained a mystery.

Heme oxygenases are ubiquitous enzymes that retain a conserved α-helical structure and canonically function in regioselective cleavage of the porphyrin macrocycle of heme to release iron12,13. HO-catalyzed heme degradation also generates carbon monoxide and a tetrapyrrole cleavage product, typically biliverdin IXα, which is further modified for downstream metabolic utilization or excretion. These reactions play key roles in heme turnover, iron acquisition, oxidative protection, and cellular signaling5,6,14. There are also reports of expanded biological roles for HO proteins that are independent of heme degradation. In humans, HO1 has been reported to translocate to the nucleus upon proteolytic processing where it modulates the activity of transcription factors through an unknown mechanism that is independent of its enzymatic activity15. HO1-mediated transcriptional changes have been implicated in cell differentiation and physiological stress responses in humans and rats16,17. Pseudo-HO enzymes that retain the HO fold but lack the conserved His ligand or heme-degrading function have also been identified in multiple organisms, but their cellular roles remain largely unknown. The best-studied example is from Arabidopsis thaliana, which encodes three active (HY1, HO3, and HO4) and one inactive (HO2) heme oxygenase homologs that all localize to the chloroplast18. AtHO2 contains an Arg in place of the conserved His and lacks detectable HO activity, and knockout studies suggest an undefined role in photomorphogenesis14. Based on these reports of non-canonical HO roles, we set out to unravel the cellular function of PfHO in P. falciparum.

We localized PfHO to the parasite apicoplast and demonstrated that conditional knockdown of PfHO disrupts parasite growth and apicoplast biogenesis. We discovered that PfHO interacts with the 35-kb apicoplast genome and requires the electropositive N-terminus, which serves as an apicoplast-targeting transit peptide but is largely retained after organelle import. Loss of PfHO resulted in a strong deficiency in apicoplast-encoded RNA but not DNA levels, suggesting a key role in expression of the apicoplast genome. Phylogenetic analyses indicated that PfHO orthologs are selectively retained by hematozoan parasites, including Babesia and Theileria, but not by other apicomplexan organisms. This study illuminates an essential role for a catalytically inactive HO-like protein in gene expression and biogenesis of the Plasmodium apicoplast.

Results

PfHO is a divergent HO homolog

Sequence analysis of PfHO provides limited insights into the origin and function of this protein. Previous studies identified low (15-20%) sequence identity between PfHO and known heme oxygenases from humans (human HO1, HuHO1), cyanobacteria (Synechocystis sp. PCC 6803 HO1, SynHO1), and plants (A. thaliana HO4, AtHO4)8,9. The N-terminal 95 residues of PfHO, which show sequence similarity only to other Plasmodium orthologs, display hydrophobic and electropositive features suggestive of sub-cellular targeting (Figure 1A), which is discussed below. Sequence homology searches with PfHO using NCBI BLAST19 and Hidden Markov Modeling tools20 revealed highest identity (65-99%) to protein orthologs in other Plasmodium species, with the next highest sequence identity (30-35%) to proteins in the hematozoan parasites, Babesia and Theileria (Figure 1 – figure supplement 1). These PfHO orthologs also lack the conserved active-site His residue and other sequence features required for enzymatic function. However, we were unable to identify convincing PfHO orthologs in apicomplexan organisms outside of blood-infecting hematozoan parasites, as previously reported21. Phylogenic analysis of HO sequences from mammals, plants, algae, insects, and parasites further highlighted the divergence of heme oxygenases between these clades of organisms (Figure 1 – figure supplement 2).

Sequence and structural homology of PfHO.

A) Sequence alignment of P. falciparum (PfHO, Q8IJS6), cyanobacterial (SynHO1, P72849), and human (HuHO1, P09601) heme oxygenase homologs (Uniprot ID). Conserved histidine ligand and distal helix residues required for catalysis in SynHO1 and HuHO1 are marked in red, and identical residues in aligned sequences are in gray. Asterisks indicate identical residues in all three sequences. The predicted N-terminal signal peptide of PfHO is underlined, electropositive residues in the PfHO leader sequence are highlighted in cyan, and the black arrow marks the putative targeting peptide processing site. Colored bars below the sequence alignment mark locations of α helices observed in crystal structures of PfHO (blue), SynHO1 (orange), and HuHO1 (grey), and the AlphaFold structural prediction for PfHO (purple). B) Structural superposition of the 2.8 Å-resolution X-ray crystal structure of apo-PfHO84–305 (blue, PDB: 8ZLD) and the 2.5 Å-resolution X-ray structure of cyanobacterial, SynHO1 (orange, PDB: 1WE1). C) Structural superposition of the proximal helix for SynHO1 active site (orange), PfHO crystal structure (blue), and the AlphaFold structural prediction of PfHO (purple). D) Top-scoring protein structures in the PDB identified by the DALI server based on structural similarity to the X-ray crystal structure of PfHO84–305. RMSD is calculated in angstroms (Å), and Z-score is a unitless parameter describing similarity, where greater value indicates higher similarity32.

Sequence homology of PfHO. (A) Sequence homologs of PfHO based on BLAST1 and HMM2 sequence-similarity searches. (B) Sequence alignment of PfHO (Q8IJS6) with SynHO1 (P72849), Human HO1 (P09601), and HO homologs in Theileria orientalis (J4C2V8) and Babesia microti (A0A1R4ACC9). Uniprot accession codes given in parentheses.

Phylogenic tree of mammalian, plant, algal, and hematozoan HOs. Nodes are annotated with bootstrap values for each branch. Ortholog groups independently predicted by OrthoMCL3 are annotated for each colored section. Only select Plasmodium HOs are included for clarity, but all other known proteins of OG6_156412 are displayed. PfHO is marked in red.

X-ray crystallographic data collection and structure refinement statistics for PfHO. Statistical values given in parentheses refer to the highest resolution bin.

Sequence and structural alignment of PfHO and plant HOs. A) Sequence alignment of PfHO with Arabidopsis thaliana (O48782) and Glycine max (C6THA4) HO1s (Uniprot ID). Heme-coordinating histidine and distal helix glycines in plant HOs are highlighted in red. B) Sequence identity of PfHO with proteins identified in our sequence homology analysis showing N-terminal targeting sequence and HO-domain separately. C) Structural alignment of the 2.8 Å-resolution PfHO crystal structure (blue, PDB: 8ZLD) with X-ray structures of A. thaliana HO1 (green, PDB: 7EQH) and G. max HO1 (yellow, PDB: 7CKA).

HO surface charge features. Electrostatic surface potential maps calculated for human HO1 (PDB 1N45), SynHO1 (PDB 1WE1), and the AlphaFold-predicted structure for PfHO (HO domain only) and contoured at ±5 kT/e. Calculations were performed using the APBS PDB2PQR online software suite4.

The low level of sequence identity to known HO enzymes and loss of heme-degrading activity by PfHO strongly suggested a repurposing of the HO scaffold for an alternative function in Plasmodium. As an initial step towards understanding this divergent function, we determined a 2.8 Å-resolution X-ray crystal structure of the HO-like domain of PfHO (residues 84-305) in its unliganded state (Figure 1B and Figure 1 – figure supplement 3), as we were unable to crystalize PfHO84–305 in the presence of bound heme. The structure obtained for apo-PfHO84–305 was highly α-helical with an overall fold expected for an HO homolog (Figure 1B and Figure 1C), including strong concordance in the positions of α-helix-forming sequences between PfHO, HuHO1, and SynHO1 (Figure 1A). Although the purified protein contained residues 84-305, electron density to support structural modeling began with residue 95 at the start of an α-helix, suggesting that residues 84-94 are disordered. A structural homology search using the DALI Protein Structure Comparison Server22 revealed strong structural similarity between PfHO and structures of HO enzymes from plants, mammals, and cyanobacteria (Figure 1D). Superposition of crystal structures for PfHO and SynHO1 revealed very similar α-helical folds, with a root-mean-square deviation (RMSD) in the positions of backbone atoms of 2.11 Å for the two structures (Figure 1B).

Despite overall structural similarity between PfHO and SynHO1, we noted several points of structural divergence beyond loss of the conserved His ligand. In structures of HO enzymes (including SynHO1), heme is sequestered within an active-site binding pocket formed by a distal helix positioned above the bound heme and a proximal helix below the heme that contains the conserved His ligand (Figure 1B). In our PfHO structure, the distal helix adopted a similar architecture to that observed in the SynHO1 structure but featured bulkier, charged Lys-Glu residues in place of the conserved Gly-Gly sequence in active HOs23 (Figure 1A). In contrast to the ordered distal helix, we were unable to resolve the structure of the C-terminal region of the proximal helix (residues 112-133) of PfHO due to weak electron density for these residues, possibly reflecting static or dynamic structural disorder in this region. We note that ordered, coiled loops were observed at the C-terminal end of the proximal helix in recent structures of plant HOs24,25 (Figure 1 – figure supplement 4). Nevertheless, disorder in the C-terminal region of the proximal helix has previously been described in a structure of human HO1 bound to synthetic 5-phenylheme (where the 5-phenyl substituent sterically disrupts proximal helix structure near the α-meso carbon) and a structure of apo rat HO1 that lacked bound heme26,27.

To model a possible structure for the proximal helix of PfHO as a basis for evaluating its electrostatic properties, we turned to a predicted AlphaFold28 structure for PfHO, which was very similar to our crystal structure (backbone RMSD of 0.57 Å). AlphaFold predicts a sharp kink in the proximal helix of PfHO that extends the position of this helix by several angstroms compared to the proximal helix in SynHO1 (Figure 1C). We also noted that the PfHO AlphaFold model predicted that residues 84-94 were unstructured, consistent with our inability to observe electron density for these residues in our X-ray data. We identified changes in the calculated electrostatic surface potential29 of PfHO84–305 between the proximal and distal helices that diminish the electropositive potential in this region compared to HuHO1 and SynHO1 (Figure 1 – figure supplement 5). The positive charge character around the heme-binding pocket in canonical HOs interacts electrostatically with propionate groups of heme and mediates HO association with electronegative electron donors (e.g., ferredoxin and cytochrome P450 reductase) required for HO activity26,30,31. Based on these observations, we conclude that PfHO is a divergent HO homolog that retains the overall HO fold but has lost key active-site and surface features that suggest a unique function independent of heme degradation.

PfHO is targeted to the apicoplast organelle

The Plasmodium-specific N-terminus of PfHO has sequence features that suggested a possible role in sub-cellular targeting. These features include a hydrophobic stretch of ∼12 residues at the N-terminus followed by electropositive sequence of ∼80 residues that are characteristic of signal and transit peptides, respectively, which direct proteins to the apicoplast organelle33,34. Analysis using the apicoplast-targeting prediction software, PlasmoAP, identified strong sequence characteristics of an apicoplast transit peptide, but its SignalP 2.0 module failed to identify a signal peptide8,34. However, more recent SignalP versions (5.0 and 6.0)35 and Phobius36 strongly predicted a signal peptide with a consensus cleavage site after N18 (Figure 1A). These features, together with known HO targeting to chloroplasts in plants18 and prior detection of PfHO in pulldown studies of apicoplast-targeted proteins37, suggested that PfHO likely targets to the apicoplast.

To determine PfHO localization in P. falciparum parasites, we stably transfected Dd2 parasites with an episome encoding full-length PfHO with a C-terminal GFP-tag. Live parasite microscopy revealed focal GFP signal that was proximal to but distinct from MitoTracker Red staining of the mitochondrion (Figure 2A and Figure 2 – figure supplement 1), which is consistent with PfHO targeting to the apicoplast. Immunofluorescence analysis (IFA) of fixed parasites indicated strong co-localization between PfHO-GFP and the apicoplast acyl carrier protein (ACP, Pf3D7_0208500) (Figure 2A and Figure 2 – figure supplement 2), providing direct evidence that PfHO targets the apicoplast. Additionally, western blot analysis of parasite lysates revealed two bands by anti-GFP staining that are suggestive of precursor and N-terminally processed forms, as typically observed for apicoplast-targeted proteins33 (Figure 2B).

PfHO localization and processing.

A) Widefield fluorescence microscopy of Dd2 parasites episomally expressing PfHO-GFP. For live imaging, parasites were stained with 25 nM Mitotracker Red and 10 nM Hoechst. For IFA, parasites were fixed and stained with anti-GFP and anti-apicoplast ACP antibodies, as well as DAPI. The average Pearson correlation coefficient (rp) of red and green channels based on all images is given. Pixel intensity plots as a function of distance along the white line in merged images are displayed graphically beside each parasite. B) Western blot of untreated or Dox/IPP-treated parasites episomally expressing PfHO-GFP and stained with anti-GFP antibody. C) Live microscopy of PfHO-GFP parasites cultured 7 days in 1 µM doxycycline (Dox) and 200 µM IPP and stained with 25 nM Mitotracker Red and 10 nM Hoechst. D) Widefield fluorescence microscopy of Dd2 parasites episomally expressing PfHO N-term-GFP and stained as in panel A. E) Western blot of parasites episomally expressing PfHO N-term-GFP and stained with anti-GFP antibody. F) Live microscopy of PfHO N-term-GFP parasites cultured 7 days in 1 µM Dox and 200 µM IPP, and stained as in panel C.

Additional widefield fluorescence microscopy of live, untreated or Dox/IPP-treated Dd2 parasites episomally expressing (A) PfHO-GFP or (B) PfHO N-term-GFP and stained with 25 nM Mitotracker Red and 10 nM Hoechst.

Additional widefield immunofluorescence microscopy of fixed Dd2 parasites episomally expressing PfHO-GFP and PfHO N-Term-GFP and stained with anti-GFP and anti-apicoplast acyl carrier protein (ACP) antibodies, and DAPI.

Uncropped western blots of untreated or Dox/IPP-treated parasites episomally expressing PfHO-GFP, stained with goat anti-GFP primary and anti-goat-IRDye800 secondary antibodies, and visualized on a Licor CLx imager.

Uncropped western blots of untreated parasites episomally expressing PfHO N-term(1-83)-GFP, stained with goat anti-GFP and mouse anti-hDHFR primary antibodies and anti-goat-IRDye800 and anti-mouse-IRDye680 secondary antibodies, then visualized on a Licor CLx imager.

To further test this conclusion, we stably disrupted the apicoplast by culturing parasites for one week in 1 µM doxycycline (Dox) with rescue by 200 µM isopentenyl pyrophosphate (IPP) to decouple parasite viability from apicoplast function. In these conditions, the apicoplast is lost and proteins targeted to the organelle accumulate in dispersed cytoplasmic foci38. As expected for an apicoplast-targeted protein, episomally expressed PfHO-GFP in Dox/IPP-treated parasites displayed a speckled constellation of dispersed fluorescent foci in each cell (Figure 2C and Figure 2 – figure supplement 1). Furthermore, western blot analysis of parasite lysates after Dox/IPP-treatment revealed only a single band by anti-GFP staining at the size of the precursor protein, providing evidence that PfHO processing depends on import into the apicoplast (Figure 2B).

To directly test whether the N-terminal leader sequence of PfHO is sufficient for apicoplast targeting, we episomally expressed the PfHO N-terminus (residues 1-83) fused to a C-terminal GFP tag in Dd2 parasites and observed a nearly identical pattern of GFP signal in live and fixed parasites compared to full-length PfHO (Figure 2D and Figure 2 – figure supplements 1 and 2), as well as both precursor and processed bands by western blot analysis (Figure 2E). In Dox/IPP-treated parasites, PfHO N-term-GFP signal appeared as dispersed fluorescent foci (Figure 2F and Figure 2 – figure supplement 1). Based on these observations, we conclude that PfHO is targeted by its N-terminal leader sequence for import into the apicoplast where it undergoes proteolytic processing. This processing and previously reported protein associations for PfHO37 suggest targeting to the apicoplast matrix. Studies described below for endogenously tagged PfHO further support this conclusion.

PfHO is essential for parasite viability and apicoplast biogenesis

To directly test PfHO essentiality in blood-stage parasites, we edited the PfHO gene to enable conditional knockdown. We first used restriction endonuclease-mediated integration39 or CRISPR/Cas9 to tag PfHO with either a C-terminal GFP-tag and the DHFR-destabilization domain40,41 in 3D7 (PM1 KO)42 parasites or a C-terminal dual hemagglutinin (HA2) tag and the glmS ribozyme43,44 in Dd2 parasites, respectively (Figure 3 – figure supplement 1). Although neither system provided substantial downregulation of PfHO expression, we used the GFP-tagged parasites to confirm apicoplast targeting and processing of endogenous PfHO. Western blot analysis of parasites expressing PfHO-GFP-DHFRDD revealed two bands by anti-GFP staining and only a single precursor protein band upon apicoplast disruption in Dox/IPP conditions (Figure 3A). Additionally, co-localization of anti-GFP and anti-ACP staining by IFA confirmed apicoplast targeting of endogenous PfHO-GFP-DHFRDD (Figure 3 – figure supplement 2). Immunogold transmission electron microscopy (TEM) of fixed parasites stained with anti-GFP and anti-ACP antibodies co-localized both proteins within a single multi-membrane compartment, as expected for targeting to the apicoplast. We noted that PfHO appeared to preferentially associate with the innermost apicoplast membrane, while ACP signal was distributed throughout the apicoplast matrix (Figure 3B and Figure 3 – figure supplement 3).

PfHO is essential for parasite viability and apicoplast maintenance.

A) Western blot of untreated or Dox/IPP-treated parasites with endogenously tagged PfHO-GFP-DHFRDD. B) Immunogold TEM of a fixed 3D7 parasite endogenously expressing PfHO-GFP-DHFRDD and stained with anti-GFP (12 nm, white arrows) and anti-apicoplast ACP (18 nm) antibodies. C) Synchronized growth assay of Dd2 parasites tagged at the PfHO locus with the aptamer/TetR-DOZI system and grown ± 1 µM aTC and ± 200 µM IPP. Data points are the average ±SD of biological triplicates. Inset: western blot analysis of PfHO expression for parasites grown 7 days ±aTC with 200 µM IPP and stained with custom anti-PfHO antibody and anti-translation elongation factor 1α (Ef1α) as loading control. Densitometry of western blot bands indicated >85% reduction in PfHO expression. D) Live microscopy of PfHO-aptamer/TetR-DOZI parasites episomally expressing apicoplast-localized GFP (PfHO N-Term-GFP) grown 5 days ±aTC with 200 µM IPP. Right: Population analysis of apicoplast morphology scored for punctate versus dispersed GFP signal in 110 total parasites from biological triplicate experiments. Statistical significance was calculated by Student’s t-test. E) Quantitative PCR analysis of the apicoplast: nuclear genome ratio for PfHO-aptamer/TetR-DOZI parasites cultured 5 days ±aTC with 200 µM IPP, based on amplification of apicoplast (SufB: Pf3D7_API04700, ClpM: Pf3D7_API03600, TufA: Pf3D7_API02900) relative to nuclear (STL: Pf3D7_0717700, I5P: Pf3D7_0802500, ADSL: Pf3D7_0206700) genes. Indicated qPCR ratios were normalized to +aTC and are the average ±SD of biological triplicates. Significance of ±aTC difference was analyzed by Student’s t-test.

Schemes for modification of the PfHO genomic locus to integrate C-terminal GFP-DHFRDD or HA2-glmS tags. A) Schematic of single-crossover strategy for tagging PfHO gene locus with C-terminal GFP-DHFRDD. 1kb sequence at 3’ of PfHO coding region (green line) was used as a probe to test for integration by Southern blot, and enzyme digestion sites with expected sizes are indicated. B) Southern blot of digested parasite DNA harvested from wildtype, polyclonal PfHO-GFP-DHFRDD, and select clonal PfHO-GFP-DHFRDD cultures. C) Schematic of single-crossover integration strategy for tagging PfHO gene locus with C-terminal HA2-glmS. Primer sites used to probe integration by genome PCR are indicated. D) Genome PCR of parasite DNA harvested from wildtype parental and PfHO-HA2-glmS cultures.

Widefield immunofluorescence microscopy of fixed 3D7 parasites endogenously expressing PfHO-GFP-DHFRDD and stained with anti-GFP and anti-apicoplast acyl carrier protein (ACP) antibodies, and DAPI. Pearson correlation coefficient (rp) of red and green channels is shown in merged images in yellow.

Additional immunogold transmission electron microscopy images of apicoplasts from fixed 3D7 parasite endogenously expressing PfHO-GFP-DHFRDD and stained with anti-GFP (12 nM, green arrows) and anti-apicoplast ACP (18 nM) antibodies.

Scheme for modification of the PfHO genomic locus to integrate the aptamer/TetR-DOZI system. A) Schematic of double-crossover integration strategy for replacing PfHO gene with cDNA encoding PfHO (Toxoplasma gondii codon bias) and RNA aptamers at the 5’ and 3’ ends of the gene. 580 bp sequence in 3’ UTR of PfHO (green line) was used as a probe to test for integration by Southern blot, and enzyme digestion sites with expected sizes marked on locus and plasmid. Primers used to probe integration by genome PCR are marked on wildtype and tagged loci. B) Genome PCR of parasite DNA harvested from wildtype parental, polyclonal PfHO-aptamer/TetR-DOZI, and select PfHO-aptamer/TetR-DOZI clonal cultures. C) Southern blot of digested parasite DNA harvested from wildtype, polyclonal PfHO-aptamer/TetR-DOZI, and select clonal PfHO-aptamer/TetR-DOZI cultures. D) Complete PfHO cDNA sequence used to replace the endogenous PfHO gene (identical encoded protein sequence).

Validation of custom PfHO antibody specificity. Western blot specificity tests of custom PfHO rabbit antibody in parasite lysates. A) Lysates from asynchronous or synchronous parasite cultures stained with 1:1,000 dilution of rabbit serum prior to inoculation with PfHO protein antigen. B) Lysates from E. coli expressing PfHO84–305 and from equal numbers of synchronous 3D7 parasites stained with 1:1,000 dilution of crude serum from the final bleed of a rabbit inoculated with PfHO protein antigen. Both blots were stained with an anti-rabbit HRP secondary antibody and visualized by chemiluminescence.

Quantitative PCR of PfHO expression ±aTC with 200 µM IPP. Transcript levels of endogenous PfHO from PfHO-aptamer/TetR-DOZI parasites grown 2 or 4 days ±aTC with 200 µM IPP. PfHO transcript abundance is normalized to the average abundance of nuclear-encoded I5P (Pf3D7_0802500), ADSL (Pf3D7_0206700), and STL (Pf3D7_0717700) transcripts, and +aTC is normalized to 100%. Normalized ratios and error bars are the average ±SD of biological triplicates.

Giemsa-stained smears of PfHO-aptamer/TetR-DOZI parasites grown in ±aTC. Times indicated are hours post-synchronization.

Additional live-parasite fluorescence microscopy images of apicoplast morphology after PfHO knockdown. PfHO-aptamer/TetR-DOZI parasites episomally expressing apicoplast-localized GFP were grown 5 days ±aTC with 200 µM IPP and stained with 10 nM Hoechst.

Uncropped western blots of parasites endogenously expressing PfHO-GFP-DHFRDD that were untreated or Dox/IPP-treated for 5 days. Membrane was stained with goat anti-GFP and custom rabbit anti-PfHO primary antibodies then anti-goat-IRDye800 and anti-rabbit IRDye680 secondary antibodies. Stained membrane was visualized on a Licor CLx imager.

Uncropped western blots of Dd2 parasites tagged with PfHO-Aptamer/TetR-DOZI grown in +aTC or -aTC/IPP conditions for 7 days, stained with rabbit anti-Ef1α and custom rabbit anti-PfHO primary antibodies and anti-rabbit-IRDye800 secondary antibody. Stained membrane was visualized on a Licor CLx imager.

Uncropped Southern blot of digested parasite DNA harvested from wildtype, polyclonal PfHO-GFP-DHFRDD, and select clonal PfHO-GFP-DHFRDD cultures.

Uncropped PCR gel of parasite DNA harvested from wildtype parental, polyclonal PfHO-HA2-glmS, and select PfHO-HA2-glmS clonal cultures.

Uncropped PCR gel of parasite DNA harvested from wildtype parental, polyclonal PfHO-aptamer/TetR-DOZI, and select PfHO-aptamer/TetR-DOZI clonal cultures.

Uncropped Southern blot of digested parasite DNA harvested from wildtype, polyclonal PfHO-aptamer/TetR-DOZI, and select clonal PfHO-aptamer/TetR-DOZI cultures.

Uncropped western blot of lysates from asynchronous or synchronous 3D7 parasites stained with 1:1,000 dilution of rabbit serum prior to inoculation with PfHO protein antigen.

Uncropped western blot of lysates from E. coli expressing PfHO84–305 and from equal numbers of synchronous 3D7 parasites stained with 1:1,000 dilution of crude serum from the final bleed of a rabbit inoculated with PfHO protein antigen.

We next used CRISPR/Cas9 to tag the PfHO gene in Dd2 parasites to encode the aptamer/TetR-DOZI system, which places protein expression under control of the non-toxic small molecule, anhydrotetracycline (aTC)45. The PfHO gene was edited to include both a single aptamer at the 5’ end and a 10x aptamer cassette at the 3’ end46,47 but without introducing an epitope tag in the encoded protein sequence. Correct integration into the PfHO locus was validated by genomic PCR and Southern blot (Figure 3 – figure supplement 4). Because the protein was untagged, we made a custom rabbit polyclonal antibody that was raised against the HO-like domain of PfHO and selectively recognized PfHO expressed in parasites and E. coli (Figure 3 – figure supplement 5). Using the aptamer-tagged parasites and this custom antibody, we performed western blot analysis to confirm detection of endogenous PfHO in +aTC conditions, including observation of precursor and processed bands. Critically, we observed that growth in -aTC conditions reduced PfHO levels by >85% (Figure 3C), indicating substantial downregulation of PfHO protein expression. PfHO mRNA levels were also selectively decreased by ∼75% upon aTC washout, consistent with prior reports that TetR-DOZI association with transcripts targets mRNA to P-bodies for degradation (Figure 3 – figure supplement 6)48–50. Because of the consistency and stringency of knockdown achieved by the PfHO-aptamer/TetR-DOZI system, all subsequent knockdown experiments were performed in this line.

By synchronous growth assay we found that PfHO knockdown in -aTC conditions resulted in a severe growth defect and widespread parasite death in the third intraerythrocytic growth cycle (Figure 3C and Figure 3 – figure supplement 7). Parasite growth was fully rescued by culture supplementation with the key apicoplast product, isopentenyl pyrophosphate (IPP)38, with IPP washout after four days in - aTC/+IPP conditions resulting in rapid parasite death (Figure 3C). These observations directly support the conclusion that PfHO is essential for blood-stage parasite viability and has a critical function within the apicoplast. To test if PfHO knockdown impacted apicoplast biogenesis, we transfected PfHO-aptamer/TetR-DOZI parasites with the PfHO N-term-GFP episome to label the apicoplast. By widefield fluorescence microscopy, we observed that live parasites in +aTC conditions had focal GFP expression. In contrast, parasites cultured five days in -aTC/+IPP conditions displayed dispersed fluorescent foci in most parasites (Figure 3D and Figure 3 – figure supplement 8). Using qPCR, we determined that parasites cultured five days in -aTC/+IPP conditions showed a dramatic reduction in apicoplast genomic DNA compared to parasites grown in +aTC conditions (Figure 3E). We conclude that PfHO function is essential for apicoplast biogenesis such that its knockdown (+IPP) results in parasite progeny that lack the intact organelle.

Electropositive transit peptide of PfHO is largely retained after apicoplast import and required for essential function

Apicoplast-targeted proteins containing bipartite N-terminal leader sequences typically undergo proteolytic cleavage that fully or mostly removes the targeting peptide upon import into the organelle33,51. Western blot analyses confirmed that PfHO is N-terminally processed (Figure 2B and Figure 3A), but we noted that the size of the mature protein was several kDa larger than the estimated size of PfHO84–305 which was previously studied as the mature HO-like domain8. Using the endogenously tagged PfHO-HA2 (glmS) line, we observed that the mature protein migrated by SDS-PAGE/western blot with an apparent molecular mass of ∼34 kDa while the HO-like domain (PfHO84–305–HA2) recombinantly expressed in E. coli migrated at ∼31 kDa (Figure 4A and Figure 3 – figure supplement 4). This observation strongly suggested that only a portion of the targeting sequence was removed upon apicoplast import and that additional N-terminal sequence beyond the HO-like domain was present in mature PfHO. Based on this approximate size difference, we estimated that ∼30-40 residues of the apicoplast-targeting sequence upstream of residue 84 were likely retained in mature PfHO.

Processing and essentiality of the PfHO N-terminus.

A) Western blot of lysates from Dd2 parasites endogenously expressing PfHO-HA2 and from E. coli recombinantly expressing PfHO84–305–HA2. B) AlphaFold structure and electrostatic surface charge predicted for PfHO33–305 corresponding to mature PfHO after apicoplast import. C) Synchronized growth assays of PfHO knockdown Dd2 parasites complemented by episomal expression of the indicated PfHO constructs in ±1 µM aTC. Growth assay data points are the average ±SD of biological triplicates.

Peptide coverage of PfHO sequence detected by mass spectrometry. A) Red residues correspond to sequence detected by tryptic digest and tandem mass spectrometry after PfHO isolation from parasites. B) List of individual peptides detected within PfHO N-term and HO-domains.

qPCR analysis of PfHO knockdown in PfHO-aptamer/TetR-DOZI parasites complemented with indicated episomes. Transcript levels of endogenous PfHO from PfHO-Aptamer/TetR-DOZI + episomal complement parasites grown in ±aTC with 200 µM IPP for 5 days. PfHO transcript abundance was normalized to the average abundance of nuclear-encoded I5P (Pf3D7_0802500), ADSL (Pf3D7_0206700), and STL (Pf3D7_0717700) transcripts, and +aTC was normalized to 100%. Normalized ratios and error bars are the average ±SD of biological triplicates.

Western blot of PfHO-aptamer/TetR-DOZI parasites complemented with episomes expressing PfHO-GFP, ACPL-HO-GFP, or PfHO N-Term-GFP. Membrane was stained with goat anti-GFP primary and anti-goat-HRP secondary antibodies and visualized by chemiluminescence. PfHO-GFP parasites were grown in ±aTC with 200 µM IPP for 5 days prior to lysis.

Sequence alignment of alveolate HO-like proteins in Plasmodium falciparum, Theileria orientalis, Babesia microti, Chromera velia, and Vitrella brassicaformis. Secondary α-helical structure, predicted by AlphaFold structural models, is underlined. Alveolate-specific N-terminal α-helix is underlined in pink.

Uncropped western blots of parasites endogenously expressing PfHO-HA2 and lysates from E. coli expressing PfHO HO-like domain (PfHO84–305–HA2) stained with rat anti-HA and custom rabbit anti-PfHO primary antibodies and anti-rabbit-IRDye680 and anti-rat-IRDye800 secondary antibodies then visualized with LICOR CLx imager. Indicated molecular masses were estimated using LICOR Image Studio software based on migration of the protein standards.

Uncropped western blot of PfHO-aptamer/TetR-DOZI parasites complemented with episomes expressing PfHO-GFP, ACPL-HO-GFP, or PfHO N-Term-GFP. Membrane was stained with goat anti-GFP primary and anti-goat-HRP secondary antibodies and visualized by chemiluminescence.

To specify the N-terminus of mature PfHO, we immunoprecipitated endogenous PfHO from parasites using the PfHO-specific antibody, fractionated the eluted sample by SDS-PAGE, transferred to PVDF membrane, and performed N-terminal protein sequencing of the Coomassie-stained band corresponding to mature PfHO. This analysis suggested an N-terminal sequence of GPLGYLNR, which corresponds to a single sequence starting at residue 33 within the electropositive transit peptide of PfHO (Figure 1A). Mass spectrometry analysis of PfHO protein purified from parasites and subjected to tryptic digest provided broad peptide coverage of PfHO sequence and identified GPLGYLNR as the most N-terminal peptide that was detected (Figure 4 – figure supplement 1). The calculated molecular mass of 35.3 kDa for PfHO33–305–HA2 is similar to the observed SDS-PAGE migration for mature PfHO-HA2 ∼34 kDa (Figure 4A). Based on these observations, we conclude that PfHO is proteolytically processed upon apicoplast import to remove part but not all of the targeting peptide and result in an N-terminus at or near Gly33 in mature PfHO.

Cleavage before Gly33 leaves ∼50 residues of the electropositive targeting peptide attached to the HO-like domain of mature PfHO. Intrinsic structural disorder is a fundamental property of apicoplast-targeting peptides52. Consistent with overall structural heterogeneity in these ∼50 residues, we were unable to crystallize recombinant PfHO33–305 that matched the mature protein, despite the presence of the structured HO-like domain. Nevertheless, we note that AlphaFold predicts that residues 57-72 of the N-terminus form an α-helix that folds across the HO-like domain of PfHO between the proximal and distal helices (Figure 4B). Additionally, the abundance of Arg and Lys residues within the retained N-terminal sequence (Figure 1A) grants a strong electropositive character to the surface of mature PfHO (Figure 4B).

To test if this retained N-terminal sequence contributes to essential PfHO function beyond a role in apicoplast targeting, we performed complementation studies using PfHO knockdown parasites. We transfected the PfHO-aptamer/TetR-DOZI parasites with episomes encoding the PfHO N-terminus fused to GFP (PfHO1–83-GFP), the HO-like domain of PfHO fused to the apicoplast ACP leader sequence (1-60)33 on its N-terminus and GFP on its C-terminus (ACPL-PfHO84–305–GFP), or full-length PfHO-GFP. We first confirmed knockdown of endogenous PfHO under -aTC conditions and proper expression and processing of the episomally expressed proteins in these parasite lines (Figure 2 and Figure 4 – figure supplements 2 and 3). Although all three proteins were correctly targeted to the apicoplast and proteolytically processed, only expression of full-length PfHO with cognate leader sequence rescued parasite growth from knockdown of endogenous PfHO (Figure 4C). We conclude that the retained N-terminal sequence of mature PfHO contributes to essential function beyond its role in apicoplast targeting.

PfHO associates with the apicoplast genome and mediates apicoplast gene expression

HO enzymes associate with a range of protein-interaction partners that depend on the organism and functional context. Known HO interactors include ferredoxin in plants and bacteria25,30, cytochrome P450 reductase in mammals12,31, and direct or indirect interactions with transcription factors that impact nuclear gene expression in mammals53–55. To identify protein-interaction partners of PfHO in parasites that might give insight into its essential role in apicoplast biogenesis, we used anti-HA immunoprecipitation (IP) to isolate endogenous PfHO-HA2 from parasites. Co-purifying proteins were identified by tryptic digest and tandem mass spectrometry (MS), then compared to protein interactors identified in negative-control samples containing HA-tagged mitochondrial proteins mACP56 or cyt c57 to filter out non-specific interactions. In two independent experiments, 509 proteins co-purified with PfHO but were not detected in pulldowns of either mitochondrial control (Figure 5 – figure supplement 1). These PfHO-specific interactors included a range of cellular proteins, including proteins targeted to the apicoplast.

PfHO interactions with proteins and DNA and impacts of its knockdown on apicoplast DNA and RNA levels.

A) Spectral counts of functionally annotated, apicoplast-targeted proteins detected in two independent anti-HA IP/MS experiments on endogenously tagged PfHO-HA2. The names of proteins in DNA/RNA metabolism are highlighted blue and proteins in translation are highlighted in red. A list of all detected proteins can be found in Source Data 1. B) Representative image showing PCR amplification of nuclear-encoded (apicoplast ACP: Pf3D7_0208500) and apicoplast-encoded (SufB: Pf3D7_API04700) genes from DNA co-purified with full-length PfHO-GFP, PfHO1–83-GFP, ACPL-PfHO84–305–GFP, or ACPL-GFP by αGFP ChIP. “Input” is total parasite DNA collected after parasite lysis and sonication, and “ChIP” is DNA eluted after αGFP IP. Densitometry quantification of 3 biological replicates is plotted on right. Statistical significance of differences between PfHO and each other construct was calculated by Student’s t-tests. C) Quantitative PCR analysis of DNA isolated from tightly synchronized PfHO-aptamer/TetR-DOZI parasites grown ±1 µM aTC with 200 µM IPP and harvested at 36 and 84 hours in biological triplicates, with normalization of Ct values averaged from three apicoplast genes (SufB, TufA, ClpM) to Ct values averaged from three nuclear (STL, I5P, ADSL) genes. Grey bars represent +aTC and red bars represent –aTC, and observed ratios are displayed as percentages. D) Quantitative RT-PCR on RNA isolated from the same parasites as in panel C to determine the normalized ratio of apicoplast transcripts (SufB, TufA, ClpM) relative to nuclear (STL, I5P, ADSL) transcripts. Significance of ±aTC differences for C and D were analyzed by Student’s t-test. E) Representative time-course showing Ct values of SufB normalized to three nuclear (STL, I5P, ADSL) genes at indicated time in PfHO-aptamer/TetR-DOZI parasites grown ±1 µM aTC with 200 µM IPP. Data points are the average ±SD of biological triplicates.

Spectral counts for proteins that co-purified with PfHO and not with either mitochondrial control in both IP/MS experiments. PfHO is marked in red, and proteins predicted to be apicoplast localized based on prior IP/MS studies are marked in green. The full list of proteins is provided in Figure 5 – source data 1.

List of apicoplast-localized proteins co-purified with PfHO in two IP/MS experiments ordered by average of spectral counts for each experiment.

Functional pathway predictions for the 65 apicoplast-localized proteins that co-purified with PfHO in two IP/MS experiments.

Quantification of parasite DNA fragment size by Agilent Bioanalyzer DNA analysis after pulse-sonication shearing of parasite lysates. Our DNA shearing method produced two populations of DNA fragment sizes, 1100-1600 bp and 150-300 bp.

Steady-state PCR amplification of additional nuclear (mACP: Pf3D7_1208300) and apicoplast (ClpM: Pf3D7_API03600) genes from DNA co-purified with indicated PfHO constructs.

Additional ChIP experiments in parasites with GFP-tagged PfHO constructs. (A) Relative abundance of apicoplast-encoded genes in DNA co-purified with the indicated GFP-tagged PfHO constructs by ChIP normalized to PfHO-GFP. B) Distribution and orientation of target genes on 35 kb apicoplast genome. Orange lines mark the location of the ∼100bp qPCR amplicon for each gene. C) Relative abundance of apicoplast-encoded genes in DNA co-purified with PfHO-GFP by ChIP normalized to input. D) Steady-state PCR amplification of nuclear-encoded (aACP: Pf3D7_0208500, GGPPS: Pf3D7_1128400) and apicoplast-encoded (SufB: Pf3D7_API04700, ClpM: Pf3D7_API03600) genes from DNA co-purified with full-length PfHO-GFP by ChIP ±PFA crosslinking.

RT-qPCR data for additional apicoplast genes. Transcript levels of apicoplast, nuclear, and mitochondrial genes were assessed in PfHO-Aptamer/TetR-DOZI parasites grown in ±aTC with 200 µM IPP for 3 days (84 hours post-synchronization). Each transcript is normalized to average of two nuclear transcripts – I5P (Pf3D7_0802500) and ADSL (Pf3D7_0206700). As an additional control, mitochondrial-encoded CytB (Pf3D7_MIT02300) transcript abundance is also measured relative to nuclear controls. Error bars represent average ±SD of independent biological triplicates.

Representative time-course showing ClpM: nuclear transcript levels at indicated times during first and second cycle of parasite growth ±aTC in the presence of IPP. Large error bars between independent replicate experiments are due to large variance in total ClpM expression. Within each experiment, ClpM transcripts are >10-fold less abundant in -aTC parasites at cycle 2 time-points 72, 84, and 90 hours. Error bars represent average ±SD of independent biological triplicates.

Uncropped PCR gel of nuclear-encoded and apicoplast-encoded genes from DNA co-purified with GFP-tagged PfHO constructs. Input (“I”) is total parasite DNA collected after parasite lysis and sonication, and ChIP (“C”) is DNA eluted from αGFP IP.

Uncropped PCR gel of nuclear-encoded and apicoplast-encoded genes from DNA co-purified with GFP-tagged PfHO constructs ±crosslinking.

Because our microscopy and biochemical studies indicated exclusive PfHO localization to the apicoplast, we focused our analysis on co-purifying proteins that were known to localize to this organelle from prior IP/MS studies37,58 (Figure 5 – figure supplement 2). Of the 65 apicoplast-targeted proteins, 37 had annotated functions, and the majority were associated with nucleic acid metabolism (e.g. GyrA/B, PREX, RAP, PKII) or protein translation (e.g., EF-G/Tu/Ts, RPS1, RPL15) pathways (Figure 5A and Figure 5 – figure supplement 3). The most highly enriched PfHO-specific interactor in both IP/MS experiments was an unannotated protein (Pf3D7_1025300) that contains a putative aspartyl protease domain and shows distant structural similarity to DNA damage-inducible protein (Ddi-1, P40087) – a ubiquitin-dependent protease associated with transcription factor processing59–61 (Figure 5 – figure supplement 2). These putative interactors are consistent with reports of non-canonical HO roles in gene expression62–64, retention by mature PfHO of an electropositive N-terminus favorable for interacting with nucleic acids (Figure 4B), and apparent PfHO localization to the membrane periphery of the apicoplast lumen (Figure 3B) where the apicoplast genome, DNA replication factors, and ribosomes associate65–67. We thus considered it most likely that PfHO had an essential function in either apicoplast genome replication or DNA-dependent gene expression.

To test the capability of PfHO to associate with the apicoplast genome, we leveraged an anti-GFP chromatin IP (ChIP) assay68,69 in parasites episomally expressing the GFP-tagged PfHO constructs tested in Figure 4C, or ACPL-GFP as a negative control. We attempted to PCR or qPCR amplify multiple nuclear-and apicoplast-encoded genes in purified ChIP and input samples that had been sonicated to shear DNA into fragments ≤2-kb in size prior to IP (Figure 5-figure supplement 4). Target nuclear and apicoplast genes were both successfully amplified in all input samples. However, only the anti-GFP pulldown from parasites expressing full-length PfHO-GFP showed robust amplification of an apicoplast-but not nuclear-encoded gene (Figure 5B and Figure 5 – figure supplements 5 and 6A). Although the portion of the PfHO N-terminus retained in mature PfHO has substantial electropositive character (Figure 1A and Figure 1B) favorable for association with DNA, this sequence in the PfHO1–83-GFP construct was not sufficient for stable pull-down of apicoplast DNA. A faint amplicon for apicoplast DNA was detected for ACPL-PfHO84– 305–GFP, but this signal was >4-fold weaker than observed for full-length PHO (Figure 5B and Figure 5 – figure supplements 5 and 6A). Based on these observations, we conclude that PfHO associates with the apicoplast genome and that DNA-binding requires both the cognate N-terminus and HO-like domain.

HO proteins in other species are reported to bind nuclear DNA53, but the sequence features and nature of those associations are unclear. We note that our ChIP-PCR experiments cannot distinguish whether PfHO pull-down with the apicoplast genome reflects direct association with DNA and/or indirect interactions mediated by other proteins. Nevertheless, the unique sequence features of the PfHO N-terminus and its requirement for DNA association may suggest a Plasmodium-specific mechanism of DNA interaction that differs from other organisms. Selective interaction of full-length PfHO with the apicoplast genome was independent of the target gene amplified by PCR (Figure 5 – figure supplement 5) or qPCR and persisted in the absence of crosslinking (Figure 5 – figure supplement 6). Our observation that 12 distinct genes spanning the apicoplast genome show similar amplification in sheared PfHO ChIP samples (Figure 5 – figure supplements 5 and 6) suggests that PfHO broadly interacts with apicoplast DNA in a sequence-independent manner akin to DNA topology regulators, gyrases, ligases, and single-strand stabilizing proteins70–72, which our IP/MS data suggest are key interactors of PfHO (see discussion below).

Our observation that full-length PfHO, containing the cognate N-terminus and HO-like domain, was concordantly required for both DNA binding (Figure 5B) and essential function (Figure 4C) suggested most simply that association with the apicoplast genome was critical to PfHO function. To test possible roles for PfHO in DNA replication and/or RNA expression, we synchronized parasites to a 5-hour window and determined the impact of PfHO knockdown on DNA and RNA abundance by qPCR and RT-qPCR, respectively, in the first and second cycles after aTC washout but before apicoplast loss and parasite death. Parasites were grown in the presence of 200 µM IPP to decouple cellular viability from apicoplast-specific defects. We observed that PfHO knockdown in -aTC conditions caused a modest ∼20% decrease in apicoplast DNA levels in second-cycle parasites relative to +aTC conditions (Figure 5C). In contrast, apicoplast RNA levels were strongly reduced upon PfHO knockdown, with a 30% reduction observed in the first cycle and nearly 90% reduction in the second cycle after aTC washout (Figure 5D). This PfHO-dependent reduction in RNA abundance was observed for all tested protein-coding and non-coding apicoplast genes spanning all of the currently known or predicted polycistronic apicoplast transcripts (Figure 5 – figure supplement 7)73–76. In contrast to its impact on apicoplast RNA, PfHO knockdown had no measurable impact on RNA transcript abundance for nuclear or mitochondrial genes (Figure 5 – figure supplement 7).

To dissect the time-course of this apicoplast-specific defect in RNA abundance, we collected samples from tightly synchronized parasites throughout the first and second growth cycles after aTC washout. In +aTC conditions, apicoplast RNA levels peaked around 36 hours for most genes, consistent with prior studies of apicoplast transcription77–79. In -aTC conditions, there was a modest decrease in RNA transcript levels in the first cycle but complete failure to increase RNA abundance in the second cycle (Figure 5E and Figure 5 – figure supplement 8). We conclude that PfHO function is essential for apicoplast gene expression and that PfHO knockdown results in a specific defect in RNA abundance that underpins higher-order defects in apicoplast biogenesis that lead to parasite death (Figure 6).

Model for essential PfHO function in apicoplast genome expression and organelle biogenesis. TP = transit peptide. Scissors represent proteolytic processing of PfHO N-terminus upon apicoplast import.

Discussion

Heme metabolism is a central cellular feature and critical therapeutic vulnerability of blood-stage malaria parasites, which have evolved unusual molecular adaptations to survive and grow within heme-rich red blood cells. Hemozoin is the dominant fate of labile heme released from large-scale hemoglobin digestion. Nevertheless, malaria parasites express a divergent and inactive heme oxygenase-like protein, whose cellular function underpinning its evolutionary retention has remained mysterious. We have elucidated an essential role for PfHO within the apicoplast organelle of P. falciparum, where it associates with the apicoplast genome and nucleic acid metabolism enzymes and is required for organelle gene expression and apicoplast biogenesis.

Molecular function of PfHO

Our study unveils that P. falciparum parasites have repurposed the HO scaffold from its canonical role in heme degradation towards a divergent function required for expression of the apicoplast genome. This essential function appears to involve direct and/or indirect association with both the apicoplast genome and a variety of DNA/RNA-metabolism enzymes. Also, both the cognate N-terminal leader sequence, most of which remains attached in the mature protein, and the HO-like domain are required for function. We hypothesize that the electropositive N-terminus (Figures 1A and 5B) may mediate direct association with the apicoplast genome, while an electronegative face on the HO-like domain opposite the heme-binding region (Figure 1 – figure supplement 5) may interact with other DNA-binding proteins (e.g., gyrases and helicases) which co-purified with PfHO in IP/MS studies (Figure 6).

The specific molecular function of PfHO that impacts RNA transcript levels in the apicoplast remains to be defined, along with the broader prokaryotic-like biochemical processes that enable transcription in the apicoplast. The apicoplast genome is organized into two polycistronic operons that each consist of roughly half the genome, orient in opposite directions, and contain duplicated rRNA genes at their 5’ ends76. No promotor sequences have been identified in the apicoplast genome, but detection of transcripts up to 15 kb73 and studies on mRNA processing sites74 suggest that full polycistronic operons are transcribed and undergo “punctuation processing” at tRNAs distributed throughout transcripts80,81. The primary subunits of a prokaryotic-like RNA polymerase complex have been identified in Plasmodium. Although associated sigma factors and other interacting proteins are proposed to exist, they have thus far remained difficult to identify73,75. We initially considered a model whereby PfHO might function as a sigma factor-like adaptor that directly binds to the multi-subunit prokaryotic RNA polymerase to mediate genomic association and transcription initiation. However, none of the core RNA polymerase subunits co-purified with PfHO in our IP/MS studies, suggesting that such a function may be unlikely.

Based on PfHO association with a variety of DNA-binding proteins and topology regulators, we consider it more likely that PfHO functions in association with other effector proteins to regulate and optimize DNA topology to enable transcriptional activity. The apicoplast-targeted gyrase A, gyrase B, DDX21 DEAD-box helicase, ligase I, the helicase domain of PREX, histone-like protein (HU)82, and single-strand DNA binding protein71 all co-purify with PfHO in our IP/MS experiments (Figure 5A). Of these proteins, DNA gyrase A83, gyrase B84, PREX85, and HU86 have been directly tested as essential apicoplast biogenesis factors in Plasmodium parasites. The relationship of most of these proteins with apicoplast transcription is unknown, but their activity in other organisms may shed light on a functional pathway in Plasmodium. Prokaryotic transcription is blocked by chemical or genetic disruption of DNA gyrase87–89, a hetero-tetramer of gyrase A and B proteins90. DNA gyrase-induced negative supercoiling locally unwinds circular DNA molecules89 to facilitate transcriptional initiation by RNA polymerase complexes87,91. DNA gyrase also relieves torsional stress formed by the procession of RNA polymerase complexes along DNA molecules and prevents stalling during transcription of long, polycistronic operons92,93. Additionally, HU coordinates prokaryotic DNA structure and supercoiling in conjunction with DNA gyrase and has been reported to regulate the spatial distribution of RNA polymerase and transcription levels in E. coli94,95. Although these proteins have exclusively been associated with apicoplast DNA replication in Plasmodium, regulation of DNA supercoiling is also a major mechanism of prokaryotic transcriptional control92,96. This biological property may have been present in ancestral plastids and inherited by the prokaryotic-like Plasmodium apicoplast.

We also noted that RNA processing and translation-associated proteins co-purified with PfHO, including ribosomal RNA (rRNA) and transfer RNA (tRNA) maturation proteins, elongation factors, and ribosomal protein subunits (Figure 5A). RNA metabolism in the Plasmodium apicoplast is sparsely understood. No RNA-degrading enzymes have been identified, and the specific functions of RNA-binding proteins remain unknown. DEAD-box DNA/RNA helicases such as DDX21 have been implicated in the removal of aberrant R-loops (DNA/RNA hybrids) during RNA transcription97–99, but Plasmodium DDX21 has also been implicated with rRNA maturation in ribosome biogenesis37. Other RNA-binding proteins that co-purified with PfHO, Nop52 and RAP, also show low-level sequence homology to ribosome assembly factors100,101. Prokaryotic ribosome assembly is a co-transcriptional process regulated by, and in close proximity to RNA transcriptional machinery102–104. Apicoplast ribosome assembly is poorly studied but appears to be similar to prokaryotic systems. Indeed, a prior IP/MS study reported that apicoplast ribosomes and ribosomal assembly complexes co-purified with RNA transcription complexes and DNA topology regulators37. PfHO pulldown with RNA-processing and protein-translation components may therefore reflect the underlying physical and temporal coupling of RNA transcription and protein translation in the apicoplast. It is also possible that PfHO contributes to other aspects of RNA metabolism that remain undefined.

Evolution of PfHO and its divergent function

Although PfHO retains the conserved α-helical structure of HO enzymes, it has strikingly low sequence similarity, implying substantial evolutionary distance from well-studied HO orthologs in humans, plants, and bacteria (Figure 1 and Figure 1 – figure supplement 2). What then is the evolutionary origin of PfHO and its non-canonical function? The Plasmodium apicoplast is thought to derive from secondary endosymbiosis through ancestral engulfment of a plastid-containing red algae that had previously engulfed a photosynthetic cyanobacterium65,105–108. Subsequent loss of photosynthesis accompanied the transition to intracellular parasitism by proto-apicomplexan ancestors109,110. HO enzymes are commonly found in photosynthetic cyanobacteria and eukaryotic chloroplasts, where they initiate biosynthesis of biliverdin and other bilin chromophores utilized in phytochrome proteins for light sensing and signaling14,111,112. Retention of PfHO in the Plasmodium apicoplast likely reflects the original presence of HO enzymes in the cyanobacterial ancestors of this organelle.

The transition from free-living algae to apicomplexan parasitism involved significant genome reduction, including loss of plastid photosynthesis and phytochrome biosynthesis pathways110. These functional reductions presumably removed the selective pressure to retain enzymatic HO activity. Indeed, the only identifiable HO homologs retained in Apicomplexa are found exclusively in hematozoan parasites such as Plasmodium, Babesia, and Theileria, and have lost the active-site His ligand like PfHO (figure 1 – figure supplement 2). Insights into the evolutionary origin of PfHO are provided by comparison to HO-like proteins in coral-symbiotic chromerid algae, the closest free-living and photosynthetic relatives to apicomplexan parasites109,110. Vitrella brassicaformis and Chromera velia both express multiple HO homologs thought to be remnants of prior endosymbioses113. These HO proteins segregate phylogenetically with either active metazoan, active plant, or inactive hematozoan HOs (figure 1 – figure supplement 2) featuring the retention or loss of the active-site His ligand, respectively. Similarly, Arabidopsis thaliana expresses four chloroplast-targeted HO homologs that include three active HO enzymes and the inactive AtHO2 that lacks the conserved His ligand but contributes to photomorphogenesis14. These observations support a model in which active HO homologs were lost along with photosynthetic machinery during the transition to apicomplexan parasitism with the exclusive retention of an inactive HO-like homolog in Hematozoa.

It remains unclear why hematozoan but no other apicomplexan parasites retained an inactive HO-like homolog. The common infection of heme-rich RBCs by hematozoans may suggest that remnant heme- or porphyrin-binding activity8 may play a role in functional regulation of these HO-like proteins within the apicoplast. Indeed, apicoplast transcription78, heme released by hemoglobin digestion114, and PfHO expression (Figure 3-figure supplement 5) all peak coincidentally around 30 hours post-infection. The affinity of PfHO for heme (Kd ∼8 µM)8 is also notably similar to the ∼2 µM concentration of labile heme estimated in the parasite cytoplasm115. Future studies can test the intriguing hypothesis that labile heme levels sensed by PfHO in the apicoplast tune functional interactions by PfHO that regulate apicoplast gene expression.

Although PfHO has diverged from canonical HO function, it retains many structural and biochemical features present in metazoan HOs which are reported to mediate transcription factor activity independent of heme degradation15–17,53,116. Thus, a role for PfHO in gene expression may be an ancient functional property of the HO scaffold that was further expanded and honed by parasites after heme-degrading activity was lost. In this regard, PfHO may be conceptually similar to other parasite proteins (e.g., mitochondrial acyl carrier protein56) that have lost canonical function but whose retention of an essential role unveils a latent secondary activity that was previously un-or under-appreciated in the shadow of the dominant primary function in well-studied orthologs from other organisms.

Implications for expanded functions of N-terminal pre-sequences beyond organelle targeting

Our discovery that PfHO requires a portion of its N-terminal transit peptide for essential function within the apicoplast expands the molecular paradigm for understanding the evolution and function of organelle-targeting leader sequences. PfHO homologs in Vitrella and Chromera also contain a predicted α-helix in their N-terminal sequences that differ from chloroplast-targeting HOs in plants (figure 4 – figure supplement 4). This unique α-helix predicted in alveolate HO-like proteins adds positive-charge to the HO-like protein that we propose mediates interaction with DNA. It is possible that the alveolate ancestors of Plasmodium expanded and functionally repurposed the targeting sequences of other plastid and/or mitochondrial proteins to provide unique organelle-specific functions that remain to be discovered. The recent report that Toxoplasma gondii parasites repurpose the cleaved leader sequence of mitochondrial cyt c1 as a stable subunit of ATP synthase supports this view117,118. Identifying adaptations in P. falciparum that diverge from human host cells can reveal novel parasite vulnerabilities that underpin the development of new therapeutic strategies.

Materials and methods

Sequence homology searches and phylogenetic analyses

We acquired Plasmodium and alveolate protein sequences from VEuPathDB.org119 and all other protein sequences from UniProt.org120 databases. The protein sequence for PfHO (Pf3D7_1011900) was analyzed by NCBI Protein BLAST19 and HMMER20 with the exclusion of Plasmodium or apicomplexan organisms to identify putative orthologs. Orthologous protein sequences and select reference HO proteins were aligned via Clustal Omega121 and analyzed using Jalview122. The multi-sequence alignment was uploaded to the IQ-TREE webserver (http://iqtree.cibiv.univie.ac.at) with ultrafast bootstrap analysis. The resulting maximum likelihood phylogenetic tree from 1000 bootstrap alignments was analyzed and displayed using FigTree (http://tree.bio.ed.ac.uk/software/figtree/).

Recombinant protein expression and purification for crystal structure determination

The gene encoding residues 84-305 of PfHO was cloned into a pET21d expression vector (Novagen) using NcoI and XhoI sites, in frame with a C-terminal His6 tag8. E. coli BL21 (DE3) cells transformed with this vector were grown in LB medium supplemented with ampicillin (50 µg/mL) (Sigma A9518) and protein expression was induced with 0.5 mM isopropyl β-D-1-thiogalactopyranoside (IPTG) (Sigma 16758) at an OD of 0.5, after which the cells were grown at 20 °C overnight. Cells were harvested by centrifugation (7,000 × g, 10 min), the pellet was resuspended in binding buffer (20 mM Tris-HCl, 500 mM NaCl, 20 mM imidazole, pH 8.5), and the cells were lysed by sonication. The cell lysate was cleared by centrifugation (40,000×g, 30 min), and the supernatant was subjected to immobilized metal affinity chromatography using a 1 mL Ni-NTA agarose column (Qiagen, 30210) equilibrated with binding buffer. The protein was eluted with a buffer containing 20 mM Tris-HCl, 500 mM NaCl, and 300 mM imidazole at pH 8.5. The eluted protein was further purified by size-exclusion chromatography employing a 26/60 Superdex75 column equilibrated with a buffer containing 20 mM HEPES and 300 mM NaCl at pH 7.5. To produce the protein containing SeMet, E. coli B834 (DE3) cells transformed with the same plasmid as above were grown in minimal media (2 g L−1 NH4Cl, 6 g L−1 KH2PO4, 17g L−1 Na2HPO4·12H2O, 1 g L−1 NaCl, 1.6 mg L−1 FeCl3, 0.5 g L−1 MgSO4·7H2O, 22 mg L−1 CaCl2·6H2O, 4 g L−1 Glucose, and 50 mg L−1 L-SeMet). Purification was identical to that of the WT protein. Protein purity was confirmed by observation of a single band at the appropriate molecular mass by Coomassie-stained SDS-PAGE.

Protein crystallization

Protein in a buffer containing 20 mM HEPES and 300 mM NaCl at pH 7.5 was subjected to crystallization trials using sitting-drop vapour diffusion using commercially available screening kits form Hampton Research in an Oryx8 system (Douglas Instruments). Protein (0.5 μL at 9.5 mg mL−1) was mixed with an equal volume of reservoir solution, and crystallization plates were maintained at 20 °C for several weeks while being examined. The crystallization solutions producing the best crystals were optimized using hanging-drop geometry in 24-well plates by mixing manually 2 µL of protein solution (5.0 mg mL−1) and an equal volume of reservoir solution. The best crystals appeared in a few days in a reservoir solution containing 0.4 M (NH4)2SO4, 0.65 M Li2SO4, and 0.1 M sodium citrate tribasic dihydrate at pH 5.6 and a temperature of 20 °C. Single crystals were mounted in nylon Cryo-Loops (Hampton Research, HR4-932) coated with Paratone (Hampton Research, HR2-862) and directly transferred to liquid nitrogen for storage.

Structural data collection and processing

Diffraction data from single crystals of WT and SeMet protein were collected in beamlines AR-NW12A and BL5A, respectively, at the Photon Factory (Tsukuba, Japan) under cryogenic conditions (100 K). Diffraction images were processed with the program MOSFLM and merged and scaled with the program SCALA or AIMLESS of the CCP4 suite. The structure of the SeMet protein was solved by the method of single anomalous diffraction using the Autosol module included in the PHENIX suite. The structure of the WT protein was determined by the molecular replacement method using the coordinates of the SeMet protein from above with the program PHASER. The models were refined with the programs REFMAC5 and built manually with COOT. Validation was carried out with PROCHECK. Data collection and structure refinement statistics are given in Figure 1 – figure supplement 3. The final structural coordinates and structure factors were deposited as RCSB Protein Data Bank entry 8ZLD.

Structural visualization and analyses

A predicted structural model for PfHO (Pf3D7_1011900) was acquired from the AlphaFold Protein Structure Database (https://alphafold.ebi.ac.uk), and published HO structures were acquired from the RCSB Protein Data Bank (https://www.rcsb.org). The AlphaFold structural model, PfHO crystal structure, and HO structures were visualized and analyzed using The PyMOL Molecular Graphics System, Version 2.5, Schrödinger, LLC. Structural superpositions were performed with the PyMOL integrated command “align” and assessed by the total number of atoms aligned and RMSD (Å). We uploaded PDB files of PfHO crystal structure and AlphaFold model to the DALI protein structural comparison server (http://ekhidna2.biocenter.helsinki.fi/dali) to identify proteins structurally related to PfHO. To determine surface charge of structures, we uploaded PDB files to APBS-PDB2PQR online software suite (https://server.poissonboltzmann.org/)123, and displayed the calculated Poisson-Boltzmann surface charge using the PyMOL APBS tool 2.1 plugin.

Parasite culturing and transfection

Plasmodium falciparum Dd2124 or 3D742 parasites were cultured in human erythrocytes obtained from Barnes-Jewish Hospital (St. Louis, MO) or the University of Utah Hospital blood bank (Salt Lake City, UT) in RPMI-1640 medium (Thermo Fisher, 23400021) supplemented with 2.5 g/L Albumax I Lipid-Rich BSA (Thermo Fisher, 11020039), 15 mg/L hypoxanthine (Sigma, H9636), 110 mg/L sodium pyruvate (Sigma, P5280), 1.19 g/L HEPES (Sigma, H4034), 2.52 g/L sodium bicarbonate (Sigma, S5761), 2 g/L glucose (Sigma, G7021), and 10 mg/L gentamicin (Invitrogen, 15750060), as previously described49. Parasites were maintained at 37°C in 90% N2/5% CO2/5% O2 or in 5% CO2/95% air. For drug-induced apicoplast-disruption experiments, parasites were cultured for ∼7 days in 1 µM doxycycline (Sigma, D9891) and 200 µM isopentenyl pyrophosphate (Isoprenoids, IPP001).

Transfections were performed in 1x cytomix containing 50-100 µg DNA by electroporation of parasite-infected RBCs in 0.2 cm cuvetes using a Bio-Rad Gene Pulser Xcell system (0.31 kV, 925 µF). Transgenic parasites were selected based on plasmid resistance cassettes encoding human DHFR125, yeast DHOD126, or blasticidin-S deaminase127 and cultured in 5 nM WR99210 (Jacobus Pharmaceutical Co.), 2 µM DSM1 (Sigma, 53330401), or 6 µM blasticidin-S (Invitrogen, R21001), respectively. After stable transfection and selection, parasites were grown in the continual presence of selection drugs, and aptamer-tagged parasites were grown in 0.5-1 µM anhydrotetrocycline (Cayman Chemicals, 10009542).

Parasite growth assays

Parasites were synchronized to the ring stage with an estimated 10-15-hour synchrony window by treatment with 5% D-sorbitol (Sigma, S7900). For aptamer-based knockdown experiments, aTC was washed out during synchronization with additional 3-5 washes in media and/or PBS. Parasite growth was monitored by plating synchronized parasites at ∼1% parasitemia and allowing culture expansion over several days with daily media changes. Parasitemia was monitored daily by flow cytometry by diluting 10 µL of each parasite culture well from each of three biological replicate samples into 200 µL of 1.0 µg/mL acridine orange (Invitrogen, A3568) in phosphate buffered saline (PBS) then analyzed on a BD FACSCelesta flow cytometry system monitoring SSC-A, FSC-A, PE-A, FITC-A, and PerCP-Cy5-5-A channels.

Cloning and episomal expression of PfHO variants in parasites

The genes encoding PfHO (Pf3D7_1011900) and apicoplast ACP (Pf3D7_0208500) were PCR amplified from P. falciparum strain 3D7 cDNA using primers with ≥20 bp homology overhangs (primers 10-15) for ligation-independent insertion into the XhoI and AvrII sites of pTEOE (human DHFR selection cassette) in frame with a C-terminal GFP tag, and with expression driven by HSP86 promoter128. Correct plasmid sequences were confirmed by Sanger sequencing, and plasmids were transfected as described above in combination with 25 µg pHTH plasmid containing piggyBac transposase to drive stable, random integration into the parasite genome129.

Parasite genome editing to enable ligand-dependent regulation of PfHO expression

We first used restriction endonuclease-mediated integration39 and single-crossover homologous recombination to tag the PfHO gene to encode a C-terminal GFP-tag fused to the DHFR-destabilization domain40,41 and a single hemagglutinin (HA) tag in 3D7 (PM1 KO)42 parasites. PCR primers 18 and 19 were used to clone the 3’ 1kb DNA sequence of the PfHO gene into the XhoI and AvrII sites of the pGDB vector40 to serve as a homology region for integration. 50 µg of this plasmid along with 50 units of MfeI restriction enzyme (NEB R3589), which cuts at a single site within the PfHO gene just upstream of the homologous sequence cloned into the donor-repair pGDB plasmid, was transfected into 3D7 PM1 KO parasites42 (which express human DHFR), as described above. Parasites were positively selected with blasticidin-S in the continuous presence of trimethoprim (Sigma, T7883), cloned by limiting dilution, and genotyped by Southern blot (Figure 3-figure supplement 1). Conditional knockdown was evaluated by washing synchronized parasites 3-5x in PBS to remove trimethoprim.

We next used CRISPR/Cas9 and single-crossover homologous recombination to tag the PfHO gene to encode a C-terminal HA2 tag fused to the glmS ribozyme43,44 in Dd2 parasites. The 1 kb homology sequence or PfHO was excised from pGDB and sub-cloned by ligation into a modified pPM2GT vector130 in which the linker-GFP sequence between the AvrII and EagI sites was replaced with a HA-HA tag and stop codon followed by the 166 bp glmS ribozyme43,44. QuikChange II site-directed mutagenesis (Agilent Technologies) was used with primers 20 and 21 to introduce silent shield mutations into the PfHO homology region for purposes of CRISPR/Cas9-based genome editing, such that the AGATGG sequence in the most 3’ exon was changed to CGGTGG. A guide RNA sequence corresponding to TGAGTAGGAAATGGAGTAGA was cloned into a modified version of the previously published pAIO CRISPR/Cas9 vector131 in which the BtgZI site was replaced with a unique HindIII site to facilitate cloning49. 50 µg each of the pPM2GT and pAIO vectors were transfected into Dd2 parasites, as described above. Parasites were positively selected by WR99210, cloned by limiting dilution, and genotyped by PCR (Figure 3 – figure supplement 1). Conditional knockdown was evaluated by adding 0-10 mM glucosamine (Sigma, G1514) to synchronized parasite cultures and evaluating protein expression and parasite growth relative to untagged parental Dd2 parasites.

CRISPR/Cas9 and double-crossover homologous recombination was used to tag the PfHO gene to encode a single RNA aptamer at the 5’ end and a 10x aptamer cassette at the 3’ end for inducible knockdown with the aptamer/TetR-DOZI system46,47. A donor plasmid was created by ligation-independent insertion of a synthetic gene (gBlock, IDT) containing PfHO cDNA (T. gondii codon bias) into the linear pSN1847L vector46, along with PCR amplified (primers 22-25) left and right homology flanks corresponding to the 5’ (426 bp immediately upstream of start codon) and 3’ (455 bp starting at position 47 after the TAA stop codon) untranslated regions of PfHO. Because the aptamer sequence contains two ATG motifs that can serve as alternate translation start sites, a viral 2A peptide sequence was introduced between the 5’ aptamer sequence and the start ATG of PfHO. This donor repair plasmid (50 µg) and the pAIO CRISPR/Cas9 vector (50 µg) with guide sequence TGAGTAGGAAATGGAGTAGA targeting the 3’ end of the endogenous PfHO gene was transfected into Dd2 parasites, as described above. No shield mutation in the donor plasmid was required due to the altered codon bias of the synthetic PfHO cDNA in that vector. Parasites were positively selected with blasticidin-S in the presence of 1 µM aTC. Integration was confirmed by Southern blot and PCR (Figure 3 – figure supplement 4). No evidence of remnant WT locus was detected in polyclonal transfectants, which were used for most experiments. Parasites were also cloned by limiting dilution.

Microscopy

Live microscopy of parasites expressing GFP-tagged proteins was performed by staining mitochondria with 25 nM MitotrackerRed CMXRos (Invitrogen, M7512) for 30 minutes and staining nuclei with 5 µg/mL Hoechst 33342 (Invitrogen, H3570) for 5-10 minutes in PBS. Stained parasites were then imaged in PBS under a coverslip on an Invitrogen EVOS M5000. Images were adjusted for brightness and contrast in FIJI with linear variations equally applied across images. Signal intensity profiles were calculated for the red and green channels respectively using the FIJI plot profile tool along a single line that transects the region of highest signal for both channels (identified on images as white line).

For immunofluorescence (IFA) experiments, parasitized red blood cells were fixed in 4% paraformaldehyde and 0.0016% glutaraldehyde for 30 minutes at 25°C, then deposited onto poly-D-lysine coated coverslips. Fixed cells were permeabilized in PBS supplemented with 0.1% Triton-X100, reduced in 0.1 mg/mL NaBH4, and blocked in 3% BSA for 30 minutes. Parasites were stained with primary antibodies: mouse anti-GFP (Invitrogen, A-11120), and rabbit anti-apicoplast ACP132 at 1:100 dilution in 1% BSA for 1 hour at 25°C, washed thrice in PBS-T (PBS with 0.1% Tween-20), stained with secondary antibodies: goat anti-mouse AF488 (Invitrogen, A-11001) and goat anti-rabbit AF647 (Invitrogen, A21244) in 1% BSA for 1 hour at 25°C, and washed thrice in PBS-T before imaging. Coverslips were mounted onto slides using ProLong Diamond Antifade Mountant with DAPI (Invitrogen, P36971) overnight at 25°C, then imaged on an Axio Imager M1 epifluorescence microscope (Carl Zeiss Microimaging Inc.) equipped with a Hamamatsu ORCA-ER digital CCD camera. Images were adjusted for brightness and contrast in FIJI with linear variations equally applied across images. Pearson correlation was automatically calculated with the FIJI Coloc2 tool on unmasked images using a point spread function of 3 pixels and 50 Costes iterations.

Immunogold transmission electron microscopy was performed (Dr. Wandy Beatty, Washington University in St. Louis) as previously described133 using endogenously tagged PfHO-GFP-DHFRDD 3D7 parasites and staining with goat anti-GFP (Abcam, ab5450) and rabbit anti-apicoplast ACP134 antibodies along with gold-conjugated anti-goat and anti-rabbit secondary antibodies.

Production and validation of custom anti-PfHO rabbit antibody

The HO domain of PfHO (84-305) was cloned into pET28 with an N-terminal His-tag and start codon, purified by Ni-NTA, cleaved by thrombin, and purified by FPLC as previously reported8. Purified protein was then injected into a rabbit for polyclonal Ab production by Cocalico Biologicals Inc. (https://www.cocalicobiologicals.com) following their standard protocol. Rabbit serum was validated for specific detection of PfHO in parasite lysates and with recombinant protein expressed in bacteria prior to use (Figure 3 – figure supplement 5).

SDS-PAGE and western blots

Parasite cultures were grown to ∼10% parasitemia in 10 mL cultures for western blots and 50-100 mL cultures for immunoprecipitation. Parasites were released from red blood cells by treatment with 0.05% saponin and subsequently lysed by sonication (20 pulses 50% duty cycle, 50% power) on a Branson microtip sonicator in TBS lysis buffer (50mM Tris pH 7.4, 150 µM NaCl, 1% v/v Triton X-100) with protease inhibitors (Invitrogen, A32955) followed by incubation at 4°C for 1 hour. Recombinantly expressed protein was obtained for western blot analysis by inducing BL21 (DE3) E. coli grown to an OD of 0.5 in LB medium with 0.5 mM IPTG at 20 °C overnight. Cells were harvested by centrifugation (5,000 RPM, 10 min), and lysed with the same methods described above. Lysates were clarified by centrifugation (14,000 RPM, 10 min) and quantified by Lowry colorimetry. 50 µg of total protein was mixed in SDS sample buffer, heated at 95C for 10 minutes, and separated by electrophoresis on 12% SDS-PAGE gels in Tris-HCl buffer. Proteins were transferred onto nitrocellulose membranes using the Bio-Rad wet-transfer system for 1 hour at 100 V, and blocked with 5% non-fat milk in TBS-T (50mM Tris pH 7.4, 150 µM NaCl, 0.5% v/v Tween-20). Membranes were stained with primary antibodies: goat anti-GFP (Abcam ab5450), rat anti-HA (Roche), mouse anti-hDHFR, rabbit anti-apicoplast ACP132, custom rabbit anti-PfHO, or rabbit anti-EF1α135 diluted 1:1000-1:2500 in blocking buffer for ≥ 18 hours at 4 °C. Samples were then washed thrice in TBS-T and stained with secondary antibodies: rabbit anti-Goat HRP (Santa Cruz, sc2768), goat anti-Rabbit HRP (Invitrogen, A27036), donkey anti-rabbit IRDye800CW (LiCor, 926-32213), donkey anti-rabbit IRDye680 (LiCor, 926-68023), or donkey anti-goat IRDye800CW (LiCor, 926-32214), diluted 1:10,000 in TBS-T for 1-2 hours at 25 °C, and again washed thrice before imaging. All blots stained with horseradish peroxidase (HRP) were developed 3 minutes with Prometheus™ ProSignal Femto ECL reagent (Genesee Scientific 20-302) and imaged on a BioRad ChemiDoc MP imaging system, and all blots stained with IRDye antibodies were imaged on a Licor Odyssey CLx system. Protein size was estimated by migration relative to the protein ladder using Licor Image Studio software v5.5.4.

N-terminal protein sequencing of PfHO

A 170 mL culture of 3D7 parasites at 3% hematocrit and ∼15% asynchronous parasitemia was harvested by centrifugation followed by release of parasites from RBCs in 5% saponin (Sigma S7900). The resulting parasite pellet was subsequently lysed in 1 mL RIPA buffer with sonication and clarified by centrifugation. PfHO was isolated by immunoprecipitation using 60 µl of affinity-purified custom anti-PfHO rabbit antibody and 400 µl of Protein A dynabeads (Invitrogen, 1001D). The beads were washed 3X in RIPA buffer, eluted with 110 µl 1X SDS sample buffer, fractionated by 10% SDS-PAGE, followed by transfer to PVDF membrane and staining by Coomassie. The band corresponding to mature PfHO was excised and subjected to N-terminal sequencing by Edman degradation at the Stanford University Protein and Nucleic Acid Facility.

Immunoprecipitation

Dd2 parasites expressing endogenously tagged PfHO with C-terminal HA-HA tag were harvested from ∼75 mL of culture by centrifugation, released from RBCs by incubating in 0.05% saponin (Sigma 84510) in PBS for 5 minutes at room temperature, and pelleted by centrifugation (5000 rpm, 30 min, 4°C). Parasites were then lysed by sonication (20 pulses 50% duty cycle, 50% power) on a Branson microtip sonicator in TBS lysis buffer (50mM Tris pH 7.4, 150 µM NaCl, 1% v/v Triton X-100) with protease inhibitors (Invitrogen A32955) followed by incubation at 4°C for 1 hour and centrifugation (14,000 rpm, 10 min), The clarified lysates were mixed with equilibrated resin from 30 µl of Pierce anti-HA-tag magnetic beads (Invitrogen 88836) and incubated for 1 hour at 4°C on a rotator. Beads were placed on a magnetic stand, supernatants were removed by aspiration, and beads were washed thrice with cold TBS-T. Bound proteins were eluted with 100 µl of 8M urea (in 100 mM Tris-HCl at pH 8.8). Proteins were precipitated by adding 100% trichloroacetic acid (Sigma 76039) to a final concentration of 20% v/v and incubated on ice for 1 hour. Proteins were then pelleted by centrifugation (13,000 rpm, 25 min, 4°C) and washed once with 500 µl of cold acetone. The protein pellets were air-dried for 30 min and stored at −20°C.

Mass spectroscopy

Protein samples isolated by anti-HA-tag IP of endogenous PfHO were reduced and alkylated using 5 mM Tris (2-carboxyethyl) phosphine and 10 mM iodoacetamide, respectively, and then enzymatically digested by sequential addition of trypsin and lys-C proteases, as previously described136. The digested peptides were desalted using Pierce C18 tips (Thermo Fisher Scientific), dried, and resuspended in 5% formic acid. Approximately 1 μg of digested peptides was loaded onto a 25-cm-long, 75-μm inner diameter fused silica capillary packed in-house with bulk ReproSil-Pur 120 C18-AQ particles, as described previously137. The 140-min water-acetonitrile gradient was delivered using a Dionex Ultimate 3,000 ultra-high performance liquid chromatography system (Thermo Fisher Scientific) at a flow rate of 200 nl/min (Buffer A: water with 3% DMSO and 0.1% formic acid, and Buffer B: acetonitrile with 3% DMSO and 0.1% formic acid). Eluted peptides were ionized by the application of distal 2.2 kV and introduced into the Orbitrap Fusion Lumos mass spectrometer (Thermo Fisher Scientific) and analyzed by tandem mass spectrometry. Data were acquired using a Data-Dependent Acquisition method consisting of a full MS1 scan (resolution = 120,000) followed by sequential MS2 scans (resolution = 15,000) for the remainder of the 3-s cycle time. Data was analyzed using the Integrated Proteomics Pipeline 2 (Integrated Proteomics Applications, San Diego, CA). Data were searched against the protein database from P. falciparum 3D7 downloaded from UniprotKB (10,826 entries) on October 2013. Tandem mass spectrometry spectra were searched using the ProLuCID algorithm followed by filtering of peptide-to-spectrum matches by DTASelect using a decoy database-estimated false discovery rate of <1%. The proteomics data are deposited in the MassIVE data repository (https://massive.ucsd.edu) under the identifier MSV000094692.

Measuring apicoplast DNA and RNA abundance in parasites

Highly synchronous parasites were obtained by sorbitol synchronization of high parasitemia cultures followed by magnet-purification of schizonts after 36-40 hours using MACS LD separation columns (Miltenyi, 130-042-901) with stringent washing. Purified schizonts were allowed to reinvade fresh RBCs for 5 hours on an orbital shaker at 100 rpm in media containing 1 µM aTC. Immediately prior to experimental plating, parasites were treated with sorbitol to ensure ≤5-hour synchrony window and washed 3-5 times in media and/or PBS to remove aTC. Times listed in growth assays are post-synchronization and reflect T=0 at the time that magnet-purified schizonts were allowed to reinvade fresh RBCs.

Highly synchronous parasites were plated in 4 mL of either +aTC or -aTC media. Parasites to be harvested in the first life cycle were cultured at 3%, second cycle at 1%, and third cycle at 0.5% starting parasitemia. 4 mL of +aTC and -aTC cultures were collected for whole DNA and RNA extraction, respectively, at 36, 84, and 132 hours, snap frozen in liquid N2, and stored at −80°C. Parasite DNA was extracted with a QIAamp DNA Blood Mini kit (Qiagen, 51106), and RNA was purified by Trizol (Invitrogen, 15596026) and phenol-chloroform isolation. We converted 1 µg of purified RNA to cDNA using a SuperScript IV VILO RT kit (Invitrogen, 11766050). Since apicoplast genes are extremely AT-rich76 and mRNA transcripts are not poly-adenylated or poly-uridylylated80, gene-specific reverse primers were used to prime the reverse transcription reactions. RT-qPCR was then used to assess the DNA/RNA abundance of four nuclear genes: STL (Pf3D7_0717700), I5P (Pf3D7_0802500), ADSL (Pf3D7_0206700), and PfHO (Pf3D7_1011900), one mitochondrial gene: CytB (Pf3D7_MIT02300), and twelve apicoplast genes: rpl-4 (Pf3D7_API01300), rpl-2 (Pf3D7_API01500), rpl-14 (Pf3D7_API02000), rps-12 (Pf3D7_API02700), EF-Tu (Pf3D7_API02900), ClpM (Pf3D7_API03600), RpoC2 (Pf3D7_API04200), RpoC1 (Pf3D7_API04300), RpoB (Pf3D7_API04400), SufB (Pf3D7_API04700), ls-rRNA (Pf3D7_API06700), and ss-rRNA (Pf3D7_API05700) (primers 28-61). Invitrogen Quantstudio Real-Time PCR systems were used to quantify abundance of DNA and cDNA using SYBR green dye and primers 28-61. The relative DNA or cDNA abundance of each apicoplast gene was normalized to the average of three nuclear-encoded genes for each sample, and -aTC was compared to +aTC by the comparative Ct method138. All qPCR experiments were performed in triplicate and data was analyzed by unpaired Student’s t-test.

PfHO chromatin immunoprecipitation (ChIP) analysis

We saponin-released 75mL of high parasitemia Dd2 cultures transfected with episomes encoding expression of PfHO-GFP, PfHO1–83-GFP, ACPL-PfHO84–305–GFP, and ACPL-GFP and crosslinked in 1% paraformaldehyde for 15 minutes at 20°C, then quenched with 125mM glycine. Crosslinked parasites were transferred into 2 mL ChIP lysis/sonication buffer (200 mM NaCl, 25 mM Tris pH 7.5, 5 mM EDTA pH 8, 1% v/v Triton X-100, 0.1% SDS w/v, 0.5% sodium deoxycholate w/v, and protease inhibitors), and sonicated for 15 cycles of 30s ON/OFF at 25% power using a microtip on a Branson sonicator, then clarified by centrifugation (14,000 rpm, 10 min, 4°C). DNA fragment size after shearing was determined by Agilent Bioanalyzer DNA analysis (University of Utah DNA Sequencing Core) (Figure 5 – figure supplement 4). We collected 200µL (10%) of the clarified, sheared lysates as “input controls” and the incubated the rest with goat anti-GFP antibody (Abcam, ab5450) at 4°C overnight. Antibody-bound protein was mixed with equilibrated resin from 25 µL protein A-conjugated dynabeads (Invitrogen, 1011D) for 1 hour at 4°C rotating, then washed in lysis/sonication buffer, 1 mg/mL salmon sperm DNA (Invitrogen, AM9680) in lysis/sonication buffer, high salt wash buffer (500 mM NaCl, 25 mM Tris pH 7.5, 2 mM EDTA pH 8, 1% v/v Triton X-100, 0.1% w/v SDS, and protease inhibitors), and Tris-EDTA buffer. Samples were eluted from protein A dynabeads by two rounds of 5-minute incubation at 65°C in 100 µL elution buffer (10 mM Tris pH 8, 1 mM EDTA, 1% w/v SDS). We increased NaCl concentration in both input control and ChIP elution samples to 200 mM and added 50 µg/mL RNAse A, then incubated overnight (at least 8 hours) at 65°C to reverse crosslinks and digest RNA. We increased EDTA concentration to 5 mM and added 2 uL of 20 mg/mL Proteinase K and digested at 60°C for 1 hour, then purified DNA using the Qiagen PCR purification kit. Purified DNA was immediately used for either steady-state PCR using primers 62-73 or qPCR amplification using protocol described above. Relative quantification of steady-state PCR bands was performed by area-under-the-curve densitometry analysis in FIJI. In qPCR experiments, amplification of each gene in ChIP DNA was normalized to amplification of the same gene in DNA purified from the input control to account for variability between parasite lines. All densitometry and qPCR experiments were performed in triplicate and statistical significance of differences between PfHO-GFP and other constructs was calculated using Student’s t-test.

eLife Assessment

This important study reveals that the malaria parasite protein PfHO, though lacking typical heme oxygenase activity, is vital for the survival of Plasmodium falciparum. Structural and localization analyses showed that PfHO is essential for apicoplast maintenance, particularly in gene expression and biogenesis, indicating a novel adaptive role for this protein in parasite biology. While the results supporting the claims of the authors are convincing, the lack of data defining a molecular understanding or mechanism of action of the protein in question limits the impact of the study.

Reviewer #1 (Public Review):

Malaria parasites detoxify free heme molecules released from digested host hemoglobins by biomineralizing them into inert hemozoin. Thus, why malaria parasites retain PfHO, a dead enzyme that loses the capacity of catabolizing heme, is an outstanding question that has puzzled researchers for more than a decade. In the current manuscript, the authors addressed this question by first solving the crystal structure of PfHO and aligning it with structures of other heme oxygenase (HO) proteins. They found that the N-terminal 95 residues of PfHO, which failed to crystalize due to their disordered nature, may serve as signal and transit peptides for PfHO subcellular localization. This was confirmed by subsequent microscopic analysis with episomally expressed PfHO-GFP and a GFP reporter fused to the first 83 residues of PfHO (PfHO N-term-GFP). To investigate the functional importance of PfHO, the authors generated an anhydrotetracycline (aTC) controlled PfHO knockdown strain. Strikingly, the parasites lacking PfHO failed to grow and lost their apicoplast. Finally, by chromatin immunoprecipitation (ChIP), quantitative PCR/RT-PCR, and growth assays, the authors showed that both the cognate N-terminus and HO-like domain were required for PfHO function as an apicoplast DNA interacting protein.

The authors systemically performed multidisciplinary approaches to address this difficult question: what is the function of this enzymatically dead PfHO? I enjoyed reading this manuscript and its thoughtful discussion. This study is not of clinical importance for antimalarial treatments but also deepens our understanding of protein function evolution. While I understand these experiments are challenging to conduct in malaria parasites, the data quality of some of the experiments could be improved. For example, most of the Western blots and Southern blots are not of high quality.

Reviewer #2 (Public Review):

Summary:

Blackwell et al. investigated the structure, localization, and physiological function of Plasmodium falciparum (Pf) heme oxygenase (HO). Pf and other malaria parasites scavenge and digest large amounts of hemoglobin from red cells for sustenance. To counter the potentially cytotoxic effects of heme, it is biomineralized into hemozoin and stored in the food vacuole. Another mechanism to counteract heme toxicity is through its enzymatic degradation via heme oxygenases. However, it was previously found by the authors that PfHO lacks the ability to catalyze heme degradation, raising the intriguing question of what the physiological function of PfHO is. In the current contribution, the authors determine that PfHO localizes to the apicoplast, determine its targeting sequence, establish the essentiality of PfHO for parasite viability, and determine that PfHO is required for proper maintenance of apicoplasts and apicoplast gene expression. In sum, the authors establish an essential physiological function for PfHO, thereby providing new insights into the role of PfHO in plasmodium metabolism.

Strengths:

The studies are rigorously conducted and the results of the experiments unambiguously support a role for PfHO as being an apicoplast-targeted protein required for parasite viability and maintenance of apicoplasts.

Weaknesses:

While the studies conducted are rigorous and support the primary conclusions, the lack of experiments probing the molecular function of PfHO limits the impact of the work. Nevertheless, the knowledge that PfHO is required for parasite viability and plays a role in the maintenance of apicoplasts is still an important advance.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment:

This important study reveals that the malaria parasite protein PfHO, though lacking typical heme oxygenase activity, is vital for the survival of Plasmodium falciparum. Structural and localization analyses showed that PfHO is essential for apicoplast maintenance, particularly in gene expression and biogenesis, indicating a novel adaptive role for this protein in parasite biology. While the results supporting the claims of the authors are convincing, the lack of data defining a molecular understanding or mechanism of action of the protein in question limits the impact of the study.

We appreciate the positive assessment. We agree that further mechanistic understanding of PfHO function remains a key future challenge. Indeed, we made extensive efforts to unravel the molecular interactions and mechanisms that underpin the critical function of PfHO. We elucidated key interactions between PfHO and the apicoplast genome, reliance of these interactions on the electropositive N-terminus, association of PfHO with DNA-binding proteins, and a specific defect in apicoplast mRNA levels upon PfHO knockdown. The major limitation we faced in further defining PfHO function is the general lack of understanding of apicoplast transcription and broader gene expression in this organelle. That limitation and the challenges to overcome it go well beyond our study and will require concerted efforts across several manuscripts (likely by multiple groups) to define the mechanistic features of apicoplast gene expression. We look forward to contributing further molecular understanding of PfHO function as broader understanding of apicoplast transcription emerges.

Public Reviews:

Reviewer #1 (Public Review):

Malaria parasites detoxify free heme molecules released from digested host hemoglobins by biomineralizing them into inert hemozoin. Thus, why malaria parasites retain PfHO, a dead enzyme that loses the capacity of catabolizing heme, is an outstanding question that has puzzled researchers for more than a decade. In the current manuscript, the authors addressed this question by first solving the crystal structure of PfHO and aligning it with structures of other heme oxygenase (HO) proteins. They found that the N-terminal 95 residues of PfHO, which failed to crystalize due to their disordered nature, may serve as signal and transit peptides for PfHO subcellular localization. This was confirmed by subsequent microscopic analysis with episomally expressed PfHO-GFP and a GFP reporter fused to the first 83 residues of PfHO (PfHO N-term-GFP). To investigate the functional importance of PfHO, the authors generated an anhydrotetracycline (aTC) controlled PfHO knockdown strain. Strikingly, the parasites lacking PfHO failed to grow and lost their apicoplast. Finally, by chromatin immunoprecipitation (ChIP), quantitative PCR/RT-PCR, and growth assays, the authors showed that both the cognate N-terminus and HO-like domain were required for PfHO function as an apicoplast DNA interacting protein.

The authors systemically performed multidisciplinary approaches to address this difficult question: what is the function of this enzymatically dead PfHO? I enjoyed reading this manuscript and its thoughtful discussion. This study is not of clinical importance for antimalarial treatments but also deepens our understanding of protein function evolution. While I understand these experiments are challenging to conduct in malaria parasites, the data quality of some of the experiments could be improved. For example, most of the Western blots and Southern blots are not of high quality.

We thank the reviewer for the positive comments but are a bit puzzled by the final statement about western and Southern blot quality. We agree that the two anti-PfHO western blots probed with custom antibody (Fig. 3- source data 2 and 8) have substantial background signal in the higher molecular mass region >75 kDa. However, we note that the critical region <50 kDa is clear in both cases and readily enables target band visualization. All other western blots probing GFP or HA epitopes are of high quality with minimal off-target background. We present two Southern blot images. We agree that the signal is somewhat faint for the Southern blot demonstrating on-target integration of the aptamer/TetR-DOZI plasmid (Fig. 3- fig. supplement 4), although we note that the correct band pattern for integration is visible. We also note that the accompanying genomic PCR data is unambiguous. The Southern blot for GFPDHFRDD incorporation into the PfHO locus (Fig. 3- fig. supplement 1) has clear signal and strongly supports on-target integration. The minor background signal in the lower left region of the image does not extend into the critical lanes nor impact interpretation of correct clonal integration.

As noted below, we have obtained a second western blot image to evaluate the decrease in PfHO protein expression in -aTC conditions. This revised image, which we now include in Fig. 3, shows clean detection of the PfHO signal in the critical molecular mass region below 40 kDa in +aTC conditions and substantial loss of this signal in -aTC conditions (relative to HSP60 loading control).

Reviewer #2 (Public Review):

Summary:

Blackwell et al. investigated the structure, localization, and physiological function of Plasmodium falciparum (Pf) heme oxygenase (HO). Pf and other malaria parasites scavenge and digest large amounts of hemoglobin from red cells for sustenance. To counter the potentially cytotoxic effects of heme, it is biomineralized into hemozoin and stored in the food vacuole. Another mechanism to counteract heme toxicity is through its enzymatic degradation via heme oxygenases. However, it was previously found by the authors that PfHO lacks the ability to catalyze heme degradation, raising the intriguing question of what the physiological function of PfHO is. In the current contribution, the authors determine that PfHO localizes to the apicoplast, determine its targeting sequence, establish the essentiality of PfHO for parasite viability, and determine that PfHO is required for proper maintenance of apicoplasts and apicoplast gene expression. In sum, the authors establish an essential physiological function for PfHO, thereby providing new insights into the role of PfHO in plasmodium metabolism.

Strengths:

The studies are rigorously conducted and the results of the experiments unambiguously support a role for PfHO as being an apicoplast-targeted protein required for parasite viability and maintenance of apicoplasts.

Weaknesses:

While the studies conducted are rigorous and support the primary conclusions, the lack of experiments probing the molecular function of PfHO limits the impact of the work. Nevertheless, the knowledge that PfHO is required for parasite viability and plays a role in the maintenance of apicoplasts is still an important advance.

We appreciate the positive assessment. We agree that further mechanistic understanding of PfHO function remains a key future challenge. Indeed, we made extensive efforts to unravel the molecular interactions and mechanisms that underpin the critical function of PfHO. We elucidated key interactions between PfHO and the apicoplast genome, reliance of these interactions on the electropositive N-terminus, association of PfHO with DNA-binding proteins, and a specific defect in apicoplast mRNA levels upon PfHO knockdown. The major limitation we faced in further defining PfHO function is the general lack of understanding of apicoplast transcription and broader gene expression. That limitation and the challenges to overcome it go well beyond our study and will require concerted efforts across several manuscripts (likely by multiple groups) to define the mechanistic features of apicoplast gene expression. We look forward to contributing further molecular understanding of PfHO function as broader understanding of apicoplast transcription emerges.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Specifically, I would like to see the expression of PfHO in the 3D7 strain and PfHOaptamer/TetR-DOZI parasites detected by PfHO antibody on the same blot. The reason is that while most of the western blots show that PfHO appears as both pro- and processed-form, Figure 3-S5B shows only the processed-form of PfHO in all life stages of 3D7. It would be interesting to find out if the processing of PfHO1 is strain/stage-specific, and whether it is regulated by heme levels. It may also be interesting to find out if the pro-form of PfHO is also functional (i.e. mutate the cleavage site).

We agree with the reviewer that Fig. 3- figure supplement 5B shows predominant detection of a single band for PfHO in untagged 3D7 parasites. In our experience, the detection of the unprocessed, pro form of PfHO can vary idiosyncratically with different experiments and cultures. In support of this variable detection of unprocessed PfHO in 3D7, we note in Fig. 3A that we detected both the unprocessed and processed forms of PfHO in a western blot of endogenously tagged PfHO-GFP-DHFRDD in 3D7 parasites with an intact apicoplast. We agree with the reviewer that future studies of stage-dependent processing of PfHO may give insights into conditions that favor or disfavor detection of the unprocessed protein.

Given prior evidence for vestigial heme binding by PfHO (Sigala et al. JBC 2012), we considered whether such heme binding might modulate PfHO expression, stability, and/or function. It is unknown if heme is present inside the apicoplast, and we currently lack evidence for heme-dependent function or expression by PfHO. Future studies can test this possible dependence.

Regarding processing and possible function of the cleaved peptide, we note that the Nterminal 18 amino acids are expected to constitute the signal peptide that is cleaved cotranslationally with import into the ER. Our data indicate that PfHO undergoes further processing upon import into the apicoplast to remove a further 15 residues. We currently have no evidence nor expectation that these additional residues contribute to PfHO function beyond targeting to the apicoplast.

I am also confused as to why the authors used rabbit anti-PfHO and rabbit anti-Ef1α on the same blot for Figure 3C, which makes it difficult to appreciate the expression changes of PfHO. Given the high non-specific background of PfHO antibody shown by other Western blots (Figure 3 - Source data 2), I would like to see a blot stained with only PfHO antibody to show that expression of PfHO has been efficiently reduced in the absence of aTC.

Bands for Ef1α (50 kDa) and untagged PfHO (~32 kDa) are readily distinguished by western blot analysis based on their distinct molecular masses and electrophoretic mobilities. We agree that staining with the anti-PfHO antibody resulted in background bands in other regions of the gel image, especially in the higher molecular mass region >75 kDa. We note that additional strong evidence for down-regulation of PfHO expression is provided in Fig. 3- figure supplement 6, which shows specific loss of PfHO mRNA transcript levels in -aTC conditions by RT-qPCR.

Nevertheless, we have followed the reviewer’s suggestion and provided a new WB image of PfHO expression ±aTC (probed only with rabbit anti-PfHO antibody) that shows strong down-regulation of PfHO protein levels in -aTC conditions, consistent with the strong growth phenotype observed. We have inserted this revised, cleaner western blot image into Fig. 3 (along with detection of HSP60 levels in replicate samples as loading control) and placed the prior image into Fig. 3- figure supplement 6. In both cases, densitometry analysis indicates an 80-85% reduction in PfHO levels in -aTC conditions.

The authors proposed that PfHO interacts with apicoplast genome DNA via the electropositive N-terminus. Interestingly, these positively charged residues are not conserved between Plasmodium, Theileria, and Babesia. I will be curious to follow the authors' future work to investigate the function of this electropositive N-terminus, possibly by comparative and mutagenesis analysis.

We agree that further molecular studies of DNA-binding determinants by PfHO and its N-terminus will be insightful.

The Quantitative RT-PCR analysis revealed that loss of PfHO specifically resulted in decreased apicoplast RNA. I wonder if the authors plan to conduct RNAseq analysis on the PfHO knockdown strain across multiple life stages, to get a clearer picture of PfHO function in malaria parasites.

Our RT-qPCR data across multiple asexual stages prior to organelle loss indicate that abundance of all apicoplast-encoded transcripts drops precipitously and uniformly upon PfHO knockdown (Fig. 5- figure supplement 7). Given the small size of the apicoplast genome and the polycistronic nature of apicoplast transcription, we assume that RNA-Seq studies would result in a similar observation. We hypothesize that PfHO knockdown and subsequent dysfunctions may interfere with RNA polymerase assembly on DNA and/or processivity. We are currently testing these hypotheses.

I noticed that the authors did not discuss the function of PfHO in apicoplast organelle biogenesis. Since ClpM (previously termed ClpC) is the only apicoplast-encoded Clp subunit that is essential for apicoplast biogenesis, does the author think that PfHO knockdown parasites lost their apicoplast due to decreased ClpM expression? If that were the case, would episomally expression or nuclear knockin of ClpM rescue PfHO deficiency in the absence of isopentenyl pyrophosphate (IPP)?

We share the reviewer’s curiosity to understand how loss of apicoplast transcripts leads to organelle dysfunction and defective IPP synthesis. We agree that ClpM function may be critical to import of nuclear-encoded proteins necessary for apicoplast function. SufB encoded on the apicoplast genome is also expected to be essential for Fe-S cluster synthesis in the apicoplast and to be required for Fe-S-dependent IPP synthesis. We have expanded the first Discussion section to address these possible connections.

Minor:

(1) None of the microscopy photos have scale bars.

We have added scale bars to all microscopy images.

(2) Multiple microscopy pictures show strange patches around the fluorescent signals (a grey square distinguishes from the black background). This is especially evident in Figure 2 S2. Was it caused by the reduction of the original pictures?

We have reviewed all fluorescence microscopy images but are unable to identify the issue noted by the reviewer. We have uploaded new versions of all images to include scale bars (as requested above), and we hope that this update resolves the issue observed by the reviewer. We are happy to further troubleshoot and address if the reviewer continues to see these artifacts and can provide further information.

(3) A description of how Southern blotting was performed is missing.

We thank the reviewer for bringing this omission to our attention. We have added a description of the Southern blot methods to the section on genome editing.

(4) Figure 3B: should be "αGFP: 12nm", not "αPfHO1: 12nm".

We have modified this labeling to read “αGFP (PfHO): 12 nm”.

(5) Figure 3C: which clone of PfHO knockdown was used in all the following figures? How many clones were tested in the following figures (did they show consistent phenotype)?

The polyclonal culture of PfHO-aptamer/TetR-DOZI knockdown parasites from transfection 11 was used for growth assay and western blot experiments, since there was no evidence by PCR or Southern blot for the wildtype PfHO locus. We have elaborated on these details in the Methods section.

Reviewer #2 (Recommendations For The Authors):

In Figure 2 and Figure 3B, to address rigor and reproducibility, the authors should state the number of parasites analyzed and if there was any variation in localization. For instance, did all of the parasites analyzed have apicoplast localization of heme oxygenase or was there a distribution of apicoplast and non-apicoplast localization?

Localization by fluorescence microscopy of episomal and endogenous tagged PfHO is presented in Fig. 2, Fig. 2- fig. supplements 1 and 2, and Fig. 3- fig. supplement 2. Localization by immunogold EM is presented in Fig. 3B and Fig. 3- fig. supplement 3. In all cases 3-4 representative images are presented that support exclusive localization of PfHO to the apicoplast. We imaged ≥10-20 additional parasites in all cases (and across distinct transfections and biological samples) that also supported exclusive localization to the apicoplast. We have modified the figure legends and methods description to note these replicate values. Finally, we note that IPP rescue of parasite viability upon PfHO knockdown strongly supports the conclusion that the critical and essential function of PfHO impacts the apicoplast, consistent with its exclusive detection in that organelle by microscopy.

Version 2

Abstract

Abstract

Malaria parasites have evolved unusual metabolic adaptations that specialize them for growth within heme-rich human erythrocytes. During blood-stage infection, Plasmodium falciparum parasites internalize and digest abundant host hemoglobin within the digestive vacuole. This massive catabolic process generates copious free heme, most of which is biomineralized into inert hemozoin. Parasites also express a divergent heme oxygenase (HO)-like protein (PfHO) that lacks key active-site residues and has lost canonical HO activity. The cellular role of this unusual protein that underpins its retention by parasites has been unknown. To unravel PfHO function, we first determined a 2.8 Å-resolution X-ray structure that revealed a highly α-helical fold indicative of distant HO homology. Localization studies unveiled PfHO targeting to the apicoplast organelle, where it is imported and undergoes N-terminal processing but retains most of the electropositive transit peptide. We observed that conditional knockdown of PfHO was lethal to parasites, which died from defective apicoplast biogenesis and impaired isoprenoid-precursor synthesis. Complementation and molecular-interaction studies revealed an essential role for the electropositive N-terminus of PfHO, which selectively associates with the apicoplast genome and enzymes involved in nucleic acid metabolism and gene expression. PfHO knockdown resulted in a specific deficiency in levels of apicoplast-encoded RNA but not DNA. These studies reveal an essential function for PfHO in apicoplast maintenance and suggest that Plasmodium repurposed the conserved HO scaffold from its canonical heme-degrading function in the ancestral chloroplast to fulfill a critical adaptive role in organelle gene expression.

Introduction

Malaria remains a devastating infectious disease marked by increasing treatment failures with frontline artemisinin therapies1,2. Plasmodium malaria parasites diverged early in eukaryotic evolution from well-studied yeast and mammalian cells, upon which most understanding of eukaryotic biology is based. Due to this evolutionary divergence, parasites acquired unusual molecular adaptations for specialized growth and survival within human erythrocytes, human hepatocytes, and mosquitos. Unraveling and understanding these adaptations will provide deep insights into the evolution of Plasmodium and other apicomplexan parasites and unveil new parasite-specific vulnerabilities that are suitable for therapeutic targeting to combat increasing parasite drug resistance.

Heme metabolism is central to parasite survival within red blood cells (RBCs), the most heme-rich cell in the human body. During blood-stage growth, Plasmodium parasites internalize and digest up to 80% of host hemoglobin within the acidic digestive vacuole (DV)3,4. This massive digestive process liberates an excess of cytotoxic free heme that is detoxified in situ within the DV via biomineralization into chemically inert hemozoin crystals. Other hematophagous organisms and cells, including certain blood-feeding insects and human liver/splenic macrophages that process senescent RBCs, depend on canonical heme oxygenase (HO) enzymes to degrade excess heme and recycle iron5–7. In contrast to these examples, Plasmodium parasites lack an active HO pathway for enzymatic heme degradation and rely fully on alternative mechanisms for heme detoxification and iron acquisition during blood-stage infection8. Nevertheless, malaria parasites express a divergent HO-like protein (PfHO, Pf3D7_1011900) with unusual biochemical features9. Although this protein shows distant homology to HO enzymes, it lacks the strictly conserved active-site His ligand and does not degrade heme in vitro or in live cells8. Genome-wide knockout and insertional mutagenesis studies in P. berghei and P. falciparum reported that the PfHO gene was refractory to disruption10,11, but the biological function that underpins the retention and putative essentiality of this divergent HO-like protein in malaria parasites has remained a mystery.

Heme oxygenases are ubiquitous enzymes that retain a conserved α-helical structure and canonically function in regioselective cleavage of the porphyrin macrocycle of heme to release iron12,13. HO-catalyzed heme degradation also generates carbon monoxide and a tetrapyrrole cleavage product, typically biliverdin IXα, which is further modified for downstream metabolic utilization or excretion. These reactions play key roles in heme turnover, iron acquisition, oxidative protection, and cellular signaling5,6,14. There are also reports of expanded biological roles for HO proteins that are independent of heme degradation. In humans, HO1 has been reported to translocate to the nucleus upon proteolytic processing where it modulates the activity of transcription factors through an unknown mechanism that is independent of its enzymatic activity15. HO1-mediated transcriptional changes have been implicated in cell differentiation and physiological stress responses in humans and rats16,17. Pseudo-HO enzymes that retain the HO fold but lack the conserved His ligand or heme-degrading function have also been identified in multiple organisms, but their cellular roles remain largely unknown. The best-studied example is from Arabidopsis thaliana, which encodes three active (HY1, HO3, and HO4) and one inactive (HO2) heme oxygenase homologs that all localize to the chloroplast18. AtHO2 contains an Arg in place of the conserved His and lacks detectable HO activity, and knockout studies suggest an undefined role in photomorphogenesis14. Based on these reports of non-canonical HO roles, we set out to unravel the cellular function of PfHO in P. falciparum.

We localized PfHO to the parasite apicoplast and demonstrated that conditional knockdown of PfHO disrupts parasite growth and apicoplast biogenesis. We discovered that PfHO interacts with the 35-kb apicoplast genome and requires the electropositive N-terminus, which serves as an apicoplast-targeting transit peptide but is largely retained after organelle import. Loss of PfHO resulted in a strong deficiency in apicoplast-encoded RNA but not DNA levels, suggesting a key role in expression of the apicoplast genome. Phylogenetic analyses indicated that PfHO orthologs are selectively retained by hematozoan parasites, including Babesia and Theileria, but not by other apicomplexan organisms. This study illuminates an essential role for a catalytically inactive HO-like protein in gene expression and biogenesis of the Plasmodium apicoplast.

Results

PfHO is a divergent HO homolog

Sequence analysis of PfHO provides limited insights into the origin and function of this protein. Previous studies identified low (15-20%) sequence identity between PfHO and known heme oxygenases from humans (human HO1, HuHO1), cyanobacteria (Synechocystis sp. PCC 6803 HO1, SynHO1), and plants (A. thaliana HO4, AtHO4)8,9. The N-terminal 95 residues of PfHO, which show sequence similarity only to other Plasmodium orthologs, display hydrophobic and electropositive features suggestive of sub-cellular targeting (Figure 1A), which is discussed below. Sequence homology searches with PfHO using NCBI BLAST19 and Hidden Markov Modeling tools20 revealed highest identity (65-99%) to protein orthologs in other Plasmodium species, with the next highest sequence identity (30-35%) to proteins in the hematozoan parasites, Babesia and Theileria (Figure 1 – figure supplement 1). These PfHO orthologs also lack the conserved active-site His residue and other sequence features required for enzymatic function. However, we were unable to identify convincing PfHO orthologs in apicomplexan organisms outside of blood-infecting hematozoan parasites, as previously reported21. Phylogenic analysis of HO sequences from mammals, plants, algae, insects, and parasites further highlighted the divergence of heme oxygenases between these clades of organisms (Figure 1 – figure supplement 2).

Sequence and structural homology of PfHO.

A) Sequence alignment of P. falciparum (PfHO, Q8IJS6), cyanobacterial (SynHO1, P72849), and human (HuHO1, P09601) heme oxygenase homologs (Uniprot ID). Conserved histidine ligand and distal helix residues required for catalysis in SynHO1 and HuHO1 are marked in red, and identical residues in aligned sequences are in gray. Asterisks indicate identical residues in all three sequences. The predicted N-terminal signal peptide of PfHO is underlined, electropositive residues in the PfHO leader sequence are highlighted in cyan, and the black arrow marks the putative targeting peptide processing site. Colored bars below the sequence alignment mark locations of α helices observed in crystal structures of PfHO (blue), SynHO1 (orange), and HuHO1 (grey), and the AlphaFold structural prediction for PfHO (purple). B) Structural superposition of the 2.8 Å-resolution X-ray crystal structure of apo-PfHO84–305 (blue, PDB: 8ZLD) and the 2.5 Å-resolution X-ray structure of cyanobacterial, SynHO1 (orange, PDB: 1WE1). C) Structural superposition of the proximal helix for SynHO1 active site (orange), PfHO crystal structure (blue), and the AlphaFold structural prediction of PfHO (purple). D) Top-scoring protein structures in the PDB identified by the DALI server based on structural similarity to the X-ray crystal structure of PfHO84–305. RMSD is calculated in angstroms (Å), and Z-score is a unitless parameter describing similarity, where greater value indicates higher similarity32.

Figure supplement 1. Sequence homologs of PfHO based on BLAST19 and HMM20 searches.

Figure supplement 2. Phylogenic tree of mammalian, plant, algal, and hematozoan HOs.

Figure supplement 3. X-ray crystallographic data collection and structure refinement statistics for PfHO.

Figure supplement 4. Sequence and structural alignments of PfHO to plant HOs (AtHO1, GmHO1).

Figure supplement 5. HO surface charge features.

Source data 1. PDB file for 2.8 Å-resolution structure of PfHO.

The low level of sequence identity to known HO enzymes and loss of heme-degrading activity by PfHO strongly suggested a repurposing of the HO scaffold for an alternative function in Plasmodium. As an initial step towards understanding this divergent function, we determined a 2.8 Å-resolution X-ray crystal structure of the HO-like domain of PfHO (residues 84-305) in its unliganded state (Figure 1B and Figure 1 – figure supplement 3), as we were unable to crystalize PfHO84–305 in the presence of bound heme. The structure obtained for apo-PfHO84–305 was highly α-helical with an overall fold expected for an HO homolog (Figure 1B and Figure 1C), including strong concordance in the positions of α-helix-forming sequences between PfHO, HuHO1, and SynHO1 (Figure 1A). Although the purified protein contained residues 84-305, electron density to support structural modeling began with residue 95 at the start of an α-helix, suggesting that residues 84-94 are disordered. A structural homology search using the DALI Protein Structure Comparison Server22 revealed strong structural similarity between PfHO and structures of HO enzymes from plants, mammals, and cyanobacteria (Figure 1D). Superposition of crystal structures for PfHO and SynHO1 revealed very similar α-helical folds, with a root-mean-square deviation (RMSD) in the positions of backbone atoms of 2.11 Å for the two structures (Figure 1B).

Despite overall structural similarity between PfHO and SynHO1, we noted several points of structural divergence beyond loss of the conserved His ligand. In structures of HO enzymes (including SynHO1), heme is sequestered within an active-site binding pocket formed by a distal helix positioned above the bound heme and a proximal helix below the heme that contains the conserved His ligand (Figure 1B). In our PfHO structure, the distal helix adopted a similar architecture to that observed in the SynHO1 structure but featured bulkier, charged Lys-Glu residues in place of the conserved Gly-Gly sequence in active HOs23 (Figure 1A). In contrast to the ordered distal helix, we were unable to resolve the structure of the C-terminal region of the proximal helix (residues 112-133) of PfHO due to weak electron density for these residues, possibly reflecting static or dynamic structural disorder in this region. We note that ordered, coiled loops were observed at the C-terminal end of the proximal helix in recent structures of plant HOs24,25 (Figure 1 – figure supplement 4). Nevertheless, disorder in the C-terminal region of the proximal helix has previously been described in a structure of human HO1 bound to synthetic 5-phenylheme (where the 5-phenyl substituent sterically disrupts proximal helix structure near the α-meso carbon) and a structure of apo rat HO1 that lacked bound heme26,27.

To model a possible structure for the proximal helix of PfHO as a basis for evaluating its electrostatic properties, we turned to a predicted AlphaFold28 structure for PfHO, which was very similar to our crystal structure (backbone RMSD of 0.57 Å). AlphaFold predicts a sharp kink in the proximal helix of PfHO that extends the position of this helix by several angstroms compared to the proximal helix in SynHO1 (Figure 1C). We also noted that the PfHO AlphaFold model predicted that residues 84-94 were unstructured, consistent with our inability to observe electron density for these residues in our X-ray data. We identified changes in the calculated electrostatic surface potential29 of PfHO84–305 between the proximal and distal helices that diminish the electropositive potential in this region compared to HuHO1 and SynHO1 (Figure 1 – figure supplement 5). The positive charge character around the heme-binding pocket in canonical HOs interacts electrostatically with the propionate groups of heme and mediates HO association with electronegative electron donors (e.g., ferredoxin and cytochrome P450 reductase) required for HO activity26,30,31. Based on these observations, we conclude that PfHO is a divergent HO homolog that retains the overall HO fold but has lost key active-site and surface features that suggest a unique function independent of heme degradation.

PfHO is targeted to the apicoplast organelle

The Plasmodium-specific N-terminus of PfHO has sequence features that suggested a possible role in sub-cellular targeting. These features include a hydrophobic stretch of ∼12 residues at the N-terminus followed by electropositive sequence of ∼80 residues that are characteristic of signal and transit peptides, respectively, which direct proteins to the apicoplast organelle33,34. Analysis using the apicoplast-targeting prediction software, PlasmoAP, identified strong sequence characteristics of an apicoplast transit peptide, but its SignalP 2.0 module failed to identify a signal peptide8,34. However, more recent SignalP versions (5.0 and 6.0)35 and Phobius36 strongly predicted a signal peptide with a consensus cleavage site after N18 (Figure 1A). These features, together with known HO targeting to chloroplasts in plants18 and prior detection of PfHO in pulldown studies of apicoplast-targeted proteins37, suggested that PfHO likely targets to the apicoplast.

To determine PfHO localization in P. falciparum parasites, we stably transfected Dd2 parasites with an episome encoding full-length PfHO with a C-terminal GFP-tag. Live parasite microscopy revealed focal GFP signal that was proximal to but distinct from MitoTracker Red staining of the mitochondrion (Figure 2A and Figure 2 – figure supplement 1), which is consistent with PfHO targeting to the apicoplast. Immunofluorescence analysis (IFA) of fixed parasites indicated strong co-localization between PfHO-GFP and the apicoplast acyl carrier protein (ACP, Pf3D7_0208500) (Figure 2A and Figure 2 – figure supplement 2), providing direct evidence that PfHO targets the apicoplast. Additionally, western blot analysis of parasite lysates revealed two bands by anti-GFP staining that are suggestive of precursor and N-terminally processed forms, as typically observed for apicoplast-targeted proteins33 (Figure 2B).

To further test this conclusion, we stably disrupted the apicoplast by culturing parasites for one week in 1 µM doxycycline (Dox) with rescue by 200 µM isopentenyl pyrophosphate (IPP) to decouple parasite viability from apicoplast function. In these conditions, the apicoplast is lost and proteins targeted to the organelle accumulate in dispersed cytoplasmic foci38. As expected for an apicoplast-targeted protein, episomally expressed PfHO-GFP in Dox/IPP-treated parasites displayed a speckled constellation of dispersed fluorescent foci in each cell (Figure 2C and Figure 2 – figure supplement 1). Furthermore, western blot analysis of parasite lysates after Dox/IPP-treatment revealed only a single band by anti-GFP staining at the size of the precursor protein, providing evidence that PfHO processing depends on import into the apicoplast (Figure 2B).

To directly test whether the N-terminal leader sequence of PfHO is sufficient for apicoplast targeting, we episomally expressed the PfHO N-terminus (residues 1-83) fused to a C-terminal GFP tag in Dd2 parasites and observed a nearly identical pattern of GFP signal in live and fixed parasites compared to full-length PfHO (Figure 2D and Figure 2 – figure supplements 1 and 2), as well as both precursor and processed bands by western blot analysis (Figure 2E). In Dox/IPP-treated parasites, PfHO N-term-GFP signal appeared as dispersed fluorescent foci (Figure 2F and Figure 2 – figure supplement 1). Based on these observations, we conclude that PfHO is targeted by its N-terminal leader sequence for import into the apicoplast where it undergoes proteolytic processing. This processing and previously reported protein associations for PfHO37 suggest targeting to the apicoplast matrix. Studies described below for endogenously tagged PfHO further support this conclusion.

PfHO is essential for parasite viability and apicoplast biogenesis

To directly test PfHO essentiality in blood-stage parasites, we edited the PfHO gene to enable conditional knockdown. We first used restriction endonuclease-mediated integration39 or CRISPR/Cas9 to tag PfHO with either a C-terminal GFP-tag and the DHFR-destabilization domain40,41 in 3D7 (PM1 KO)42 parasites or a C-terminal dual hemagglutinin (HA2) tag and the glmS ribozyme43,44 in Dd2 parasites, respectively (Figure 3 – figure supplement 1). Although neither system provided substantial downregulation of PfHO expression, we used the GFP-tagged parasites to confirm apicoplast targeting and processing of endogenous PfHO. Western blot analysis of parasites expressing PfHO-GFP-DHFRDD revealed two bands by anti-GFP staining and only a single precursor protein band upon apicoplast disruption in Dox/IPP conditions (Figure 3A). Additionally, co-localization of anti-GFP and anti-ACP staining by IFA confirmed apicoplast targeting of endogenous PfHO-GFP-DHFRDD (Figure 3 – figure supplement 2). Immunogold transmission electron microscopy (TEM) of fixed parasites stained with anti-GFP and anti-ACP antibodies co-localized both proteins within a single multi-membrane compartment, as expected for targeting to the apicoplast. We noted that PfHO appeared to preferentially associate with the innermost apicoplast membrane, while ACP signal was distributed throughout the apicoplast matrix (Figure 3B and Figure 3 – figure supplement 3).

PfHO localization and processing.

A) Widefield fluorescence microscopy of Dd2 parasites episomally expressing PfHO-GFP. For live imaging, parasites were stained with 25 nM Mitotracker Red and 10 nM Hoechst. For IFA, parasites were fixed and stained with anti-GFP and anti-apicoplast ACP antibodies, as well as DAPI. For all images, the white scale bars indicate 1 µm. The average Pearson correlation coefficient (rp) of red and green channels based on all images is given. Pixel intensity plots as a function of distance along the white line in merged images are displayed graphically beside each parasite. B) Western blot of untreated or Dox/IPP-treated parasites episomally expressing PfHO-GFP and stained with anti-GFP antibody. C) Live microscopy of PfHO-GFP parasites cultured 7 days in 1 µM doxycycline (Dox) and 200 µM IPP and stained with 25 nM Mitotracker Red and 10 nM Hoechst. D) Widefield fluorescence microscopy of Dd2 parasites episomally expressing PfHO N-term-GFP and stained as in panel A. E) Western blot of parasites episomally expressing PfHO N-term-GFP and stained with anti-GFP antibody. F) Live microscopy of PfHO N-term-GFP parasites cultured 7 days in 1 µM Dox and 200 µM IPP, and stained as in panel C. For each parasite line and condition, ≥20 parasites were analyzed by live imaging and ≥10 parasites were analyzed by IFA.

Figure supplement 1. Additional widefield fluorescence microscopy of live Dd2 parasites episomally expressing PfHO-GFP and PfHO N-Term-GFP.

Figure supplement 2. Additional widefield IFA microscopy of fixed Dd2 parasites episomally expressing PfHO-GFP and PfHO N-Term-GFP.

Source data 1. Uncropped western blots of untreated or Dox/IPP-treated parasites episomally expressing PfHO-GFP in figure 2B.

Source data 2. Uncropped western blots of untreated parasites episomally expressing PfHO N-term-GFP in figure 2E.

PfHO is essential for parasite viability and apicoplast maintenance.

A) Western blot of untreated or Dox/IPP-treated parasites with endogenously tagged PfHO-GFP-DHFRDD. B) Immunogold TEM of a fixed 3D7 parasite endogenously expressing PfHO-GFP-DHFRDD and stained with anti-GFP (12 nm, white arrows) and anti-apicoplast ACP (18 nm) antibodies. C) Synchronized growth assay of Dd2 parasites tagged at the PfHO locus with the aptamer/TetR-DOZI system and grown ± 1 µM aTC and ± 200 µM IPP. Data points are the average ±SD of biological triplicates. Inset: western blot analysis of PfHO expression for 100 µg total lysates from parasites grown 3 days ±aTC, analyzed in duplicate samples run on the same gel, and stained with either custom anti-PfHO antibody or anti-heat shock protein 60 (HSP60) as loading control. Densitometry of western blot bands indicated >80% reduction in PfHO expression. D) Live microscopy of PfHO-aptamer/TetR-DOZI parasites episomally expressing apicoplast-localized GFP (PfHO N-Term-GFP) grown 5 days ±aTC with 200 µM IPP. White scale bars in bottom right corners are 1 µm. Right: Population analysis of apicoplast morphology scored for punctate versus dispersed GFP signal in 110 total parasites from biological triplicate experiments. Statistical significance was calculated by Student’s t-test. E) Quantitative PCR analysis of the apicoplast: nuclear genome ratio for PfHO-aptamer/TetR-DOZI parasites cultured 5 days ±aTC with 200 µM IPP, based on amplification of apicoplast (SufB: Pf3D7_API04700, ClpM: Pf3D7_API03600, TufA: Pf3D7_API02900) relative to nuclear (STL: Pf3D7_0717700, I5P: Pf3D7_0802500, ADSL: Pf3D7_0206700) genes. Indicated qPCR ratios were normalized to +aTC and are the average ±SD of biological triplicates. Significance of ±aTC difference was analyzed by Student’s t-test.

Figure supplement 1. Schemes for modification of the PfHO genomic locus to integrate the C-terminal GFP-DHFRDD or HA2-glmS tags.

Figure supplement 2. IFA microscopy of 3D7 parasites expressing PfHO-GFP-DHFRDD.

Figure supplement 3. Additional immunogold TEM images of 3D7 parasites expressing PfHO-GFP-DHFRDD.

Figure supplement 4. Scheme for modification of the PfHO genomic locus to integrate the aptamer/TetR-DOZI system.

Figure supplement 5. Validation of custom PfHO antibody specificity.

Figure supplement 6. Quantitative PCR and additional western blot analysis of PfHO expression ±aTC with 200 µM IPP.

Figure supplement 7. Giemsa-stained smears of PfHO-aptamer/TetR-DOZI parasites grown in ±aTC.

Figure supplement 8. Additional fluorescence microscopy images of PfHO-aptamer/TetR-DOZI parasites episomally expressing apicoplast-localized GFP grown 5 days ±aTC with 200 µM IPP.

Source data 1. Uncropped western blots of parasites with endogenously tagged PfHO-GFP-DHFRDD.

Source data 2. Uncropped western blots of PfHO expression.

Source data 3. Uncropped Southern blot of parasite DNA from PfHO-GFP-DHFRDD cultures.

Source data 4. Uncropped PCR gel of parasite DNA from PfHO-HA2-glmS cultures.

Source data 5. Uncropped PCR gel of parasite DNA from PfHO-aptamer/TetR-DOZI cultures.

Source data 6. Uncropped Southern blot of parasite DNA from PfHO-aptamer/TetR-DOZI cultures.

Source Data 7. Uncropped western blot of 3D7 parasites stained with rabbit serum prior to inoculation with PfHO protein antigen.

Source Data 8. Uncropped western blot of E. coli expressing PfHO84–305 and 3D7 parasites stained with crude serum from the final bleed of a rabbit inoculated with PfHO protein antigen.

We next used CRISPR/Cas9 to tag the PfHO gene in Dd2 parasites to encode the aptamer/TetR-DOZI system, which places protein expression under control of the non-toxic small molecule, anhydrotetracycline (aTC)45. The PfHO gene was edited to include both a single aptamer at the 5’ end and a 10x aptamer cassette at the 3’ end46,47 but without introducing an epitope tag in the encoded protein sequence. Correct integration into the PfHO locus was validated by genomic PCR and Southern blot (Figure 3 – figure supplement 4). Because the protein was untagged, we made a custom rabbit polyclonal antibody that was raised against the HO-like domain of PfHO and selectively recognized PfHO expressed in parasites and E. coli (Figure 3 – figure supplement 5). Using the aptamer-tagged parasites and this custom antibody, we performed western blot analysis to confirm detection of endogenous PfHO in +aTC conditions, including observation of precursor and processed bands. Critically, we observed that growth in -aTC conditions reduced PfHO levels by ≥80% across biological replicate samples (Figure 3C and Figure 3 – figure supplement 6), indicating substantial downregulation of PfHO protein expression. PfHO mRNA levels were also selectively decreased by ∼75% upon aTC washout, consistent with prior reports that TetR-DOZI association with transcripts targets mRNA to P-bodies for degradation (Figure 3 – figure supplement 6)48–50. Because of the consistency and stringency of knockdown achieved by the PfHO-aptamer/TetR-DOZI system, all subsequent knockdown experiments were performed in this line.

By synchronous growth assay we found that PfHO knockdown in -aTC conditions resulted in a severe growth defect and widespread parasite death in the third intraerythrocytic growth cycle (Figure 3C and Figure 3 – figure supplement 7). Parasite growth was fully rescued by culture supplementation with the key apicoplast product, isopentenyl pyrophosphate (IPP)38, with IPP washout after four days in - aTC/+IPP conditions resulting in rapid parasite death (Figure 3C). These observations directly support the conclusion that PfHO is essential for blood-stage parasite viability and has a critical function within the apicoplast. To test if PfHO knockdown impacted apicoplast biogenesis, we transfected PfHO-aptamer/TetR-DOZI parasites with the PfHO N-term-GFP episome to label the apicoplast. By widefield fluorescence microscopy, we observed that live parasites in +aTC conditions had focal GFP expression. In contrast, parasites cultured five days in -aTC/+IPP conditions displayed dispersed fluorescent foci in most parasites (Figure 3D and Figure 3 – figure supplement 8). Using qPCR, we determined that parasites cultured five days in -aTC/+IPP conditions showed a dramatic reduction in apicoplast genomic DNA compared to parasites grown in +aTC conditions (Figure 3E). We conclude that PfHO function is essential for apicoplast biogenesis such that its knockdown (+IPP) results in parasite progeny that lack the intact organelle.

Electropositive transit peptide of PfHO is largely retained after apicoplast import and required for essential function

Apicoplast-targeted proteins containing bipartite N-terminal leader sequences typically undergo proteolytic cleavage that fully or mostly removes the targeting peptide upon import into the organelle33,51. Western blot analyses confirmed that PfHO is N-terminally processed (Figure 2B and Figure 3A), but we noted that the size of the mature protein was several kDa larger than the estimated size of PfHO84–305 which was previously studied as the mature HO-like domain8. Using the endogenously tagged PfHO-HA2 (glmS) line, we observed that the mature protein migrated by SDS-PAGE/western blot with an apparent molecular mass of ∼34 kDa while the HO-like domain (PfHO84–305–HA2) recombinantly expressed in E. coli migrated at ∼31 kDa (Figure 4A and Figure 3 – figure supplement 4). This observation strongly suggested that only a portion of the targeting sequence was removed upon apicoplast import and that additional N-terminal sequence beyond the HO-like domain was present in mature PfHO. Based on this approximate size difference, we estimated that ∼30-40 residues of the apicoplast-targeting sequence upstream of residue 84 were likely retained in mature PfHO.

Processing and essentiality of the PfHO N-terminus.

A) Western blot of lysates from Dd2 parasites endogenously expressing PfHO-HA2 and from E. coli recombinantly expressing PfHO84–305–HA2. B) AlphaFold structure and electrostatic surface charge predicted for PfHO33–305 corresponding to mature PfHO after apicoplast import. C) Synchronized growth assays of PfHO knockdown Dd2 parasites complemented by episomal expression of the indicated PfHO constructs in ±1 µM aTC. Growth assay data points are the average ±SD of biological triplicates.

Figure supplement 1. Peptide coverage of PfHO sequence detected by mass spectrometry.

Figure supplement 2. RT-qPCR analysis of endogenous PfHO transcript levels ±aTC in PfHO-aptamer/TetR-DOZI parasites complemented with PfHO episomes.

Figure supplement 3. Western blot of episomal PfHO expression in PfHO-aptamer/TetR-DOZI parasites.

Figure supplement 4. Sequence alignment of alveolate HO-like proteins indicating α-helical structure.

Source data 1. Uncropped western blots of parasites endogenously expressing PfHO-HA2 and E.coli expressing PfHO HO-like domain (PfHO84–305–HA2).

Source data 2. Uncropped western blot of episomal PfHO expression in PfHO-aptamer/TetR-DOZI parasites.

To specify the N-terminus of mature PfHO, we immunoprecipitated endogenous PfHO from parasites using the PfHO-specific antibody, fractionated the eluted sample by SDS-PAGE, transferred to PVDF membrane, and performed N-terminal protein sequencing of the Coomassie-stained band corresponding to mature PfHO. This analysis suggested an N-terminal sequence of GPLGYLNR, which corresponds to a single sequence starting at residue 33 within the electropositive transit peptide of PfHO (Figure 1A). Mass spectrometry analysis of PfHO protein purified from parasites and subjected to tryptic digest provided broad peptide coverage of PfHO sequence and identified GPLGYLNR as the most N-terminal peptide that was detected (Figure 4 – figure supplement 1). The calculated molecular mass of 35.3 kDa for PfHO33–305–HA2 is similar to the observed SDS-PAGE migration for mature PfHO-HA2 ∼34 kDa (Figure 4A). Based on these observations, we conclude that PfHO is proteolytically processed upon apicoplast import to remove part but not all of the targeting peptide and result in an N-terminus at or near Gly33 in mature PfHO.

Cleavage before Gly33 leaves ∼50 residues of the electropositive targeting peptide attached to the HO-like domain of mature PfHO. Intrinsic structural disorder is a fundamental property of apicoplast-targeting peptides52. Consistent with overall structural heterogeneity in these ∼50 residues, we were unable to crystallize recombinant PfHO33–305 that matched the mature protein, despite the presence of the structured HO-like domain. Nevertheless, we note that AlphaFold predicts that residues 57-72 of the N-terminus form an α-helix that folds across the HO-like domain of PfHO between the proximal and distal helices (Figure 4B). Additionally, the abundance of Arg and Lys residues within the retained N-terminal sequence (Figure 1A) grants a strong electropositive character to the surface of mature PfHO (Figure 4B).

To test if this retained N-terminal sequence contributes to essential PfHO function beyond a role in apicoplast targeting, we performed complementation studies using PfHO knockdown parasites. We transfected the PfHO-aptamer/TetR-DOZI parasites with episomes encoding the PfHO N-terminus fused to GFP (PfHO1–83-GFP), the HO-like domain of PfHO fused to the apicoplast ACP leader sequence (1-60)33 on its N-terminus and GFP on its C-terminus (ACPL-PfHO84–305–GFP), or full-length PfHO-GFP. We first confirmed knockdown of endogenous PfHO under -aTC conditions and proper expression and processing of the episomally expressed proteins in these parasite lines (Figure 2 and Figure 4 – figure supplements 2 and 3). Although all three proteins were correctly targeted to the apicoplast and proteolytically processed, only expression of full-length PfHO with cognate leader sequence rescued parasite growth from knockdown of endogenous PfHO (Figure 4C). We conclude that the retained N-terminal sequence of mature PfHO contributes to essential function beyond its role in apicoplast targeting.

PfHO associates with the apicoplast genome and mediates apicoplast gene expression

HO enzymes associate with a range of protein-interaction partners that depend on the organism and functional context. Known HO interactors include ferredoxin in plants and bacteria25,30, cytochrome P450 reductase in mammals12,31, and direct or indirect interactions with transcription factors that impact nuclear gene expression in mammals53–55. To identify protein-interaction partners of PfHO in parasites that might give insight into its essential role in apicoplast biogenesis, we used anti-HA immunoprecipitation (IP) to isolate endogenous PfHO-HA2 from parasites. Co-purifying proteins were identified by tryptic digest and tandem mass spectrometry (MS), then compared to protein interactors identified in negative-control samples containing HA-tagged mitochondrial proteins mACP56 or cyt c57 to filter out non-specific interactions. In two independent experiments, 509 proteins co-purified with PfHO but were not detected in pulldowns of either mitochondrial control (Figure 5 – figure supplement 1). These PfHO-specific interactors included a range of cellular proteins, including proteins targeted to the apicoplast.

Because our microscopy and biochemical studies indicated exclusive PfHO localization to the apicoplast, we focused our analysis on co-purifying proteins that were known to localize to this organelle from prior IP/MS studies37,58 (Figure 5 – figure supplement 2). Of the 65 apicoplast-targeted proteins, 37 had annotated functions, and the majority were associated with nucleic acid metabolism (e.g. GyrA/B, PREX, RAP, PKII) or protein translation (e.g., EF-G/Tu/Ts, RPS1, RPL15) pathways (Figure 5A and Figure 5 – figure supplement 3). The most highly enriched PfHO-specific interactor in both IP/MS experiments was an unannotated protein (Pf3D7_1025300) that contains a putative aspartyl protease domain and shows distant structural similarity to DNA damage-inducible protein (Ddi-1, P40087) – a ubiquitin-dependent protease associated with transcription factor processing59–61 (Figure 5 – figure supplement 2). These putative interactors are consistent with reports of non-canonical HO roles in gene expression62–64, retention by mature PfHO of an electropositive N-terminus favorable for interacting with nucleic acids (Figure 4B), and apparent PfHO localization to the membrane periphery of the apicoplast lumen (Figure 3B) where the apicoplast genome, DNA replication factors, and ribosomes associate65–67. We thus considered it most likely that PfHO had an essential function in either apicoplast genome replication or DNA-dependent gene expression.

PfHO interactions with proteins and DNA and impacts of its knockdown on apicoplast DNA and RNA levels.

A) Spectral counts of functionally annotated, apicoplast-targeted proteins detected in two independent anti-HA IP/MS experiments on endogenously tagged PfHO-HA2. The names of proteins in DNA/RNA metabolism are highlighted blue and proteins in translation are highlighted in red. A list of all detected proteins can be found in Source Data 1. B) Representative image showing PCR amplification of nuclear-encoded (apicoplast ACP: Pf3D7_0208500) and apicoplast-encoded (SufB: Pf3D7_API04700) genes from DNA co-purified with full-length PfHO-GFP, PfHO1–83-GFP, ACPL-PfHO84–305–GFP, or ACPL-GFP by αGFP ChIP. “Input” is total parasite DNA collected after parasite lysis and sonication, and “ChIP” is DNA eluted after αGFP IP. Densitometry quantification of 3 biological replicates is plotted on right. Statistical significance of differences between PfHO and each other construct was calculated by Student’s t-tests. C) Quantitative PCR analysis of DNA isolated from tightly synchronized PfHO-aptamer/TetR-DOZI parasites grown ±1 µM aTC with 200 µM IPP and harvested at 36 and 84 hours in biological triplicates, with normalization of Ct values averaged from three apicoplast genes (SufB, TufA, ClpM) to Ct values averaged from three nuclear (STL, I5P, ADSL) genes. Grey bars represent +aTC and red bars represent –aTC, and observed ratios are displayed as percentages. D) Quantitative RT-PCR on RNA isolated from the same parasites as in panel C to determine the normalized ratio of apicoplast transcripts (SufB, TufA, ClpM) relative to nuclear (STL, I5P, ADSL) transcripts. Significance of ±aTC differences for C and D were analyzed by Student’s t-test. E) Representative time-course showing Ct values of SufB normalized to three nuclear (STL, I5P, ADSL) genes at indicated time in PfHO-aptamer/TetR-DOZI parasites grown ±1 µM aTC with 200 µM IPP. Data points are the average ±SD of biological triplicates.

Figure supplement 1. Spectral counts for proteins co-purified with PfHO in IP/MS experiments.

Figure supplement 2. List of apicoplast-localized proteins that co-purified with PfHO in IP/MS experiments.

Figure supplement 3. Functional pathway predictions for apicoplast-localized proteins that co-purified with PfHO in IP/MS experiments.

Figure supplement 4. Fragment analyzer quantification of parasite DNA fragment size after shearing.

Figure supplement 5. Steady-state PCR amplification of additional nuclear and apicoplast genes from DNA co-purified with indicated PfHO constructs.

Figure supplement 6. Additional ChIP experiments in parasites with GFP-tagged PfHO constructs.

Figure supplement 7. RT-qPCR of additional apicoplast genes in PfHO-aptamer/TetR-DOZI parasites grown for 3 days ±1 µM aTC with 200 µM IPP.

Figure supplement 8. Representative time-course of ClpM: nuclear transcript levels in synchronous PfHO-aptamer/TetR-DOZI parasites grown ±1 µM aTC with 200 µM IPP.

Source data 1. Table of proteins identified in PfHO IP/MS experiments.

Source data 2. Uncropped PCR gel of PfHO ChIP experiments in parasites with GFP-tagged PfHO constructs

Source data 3. Uncropped PCR gel of PfHO ChIP experiments ±crosslinking

To test the capability of PfHO to associate with the apicoplast genome, we leveraged an anti-GFP chromatin IP (ChIP) assay68,69 in parasites episomally expressing the GFP-tagged PfHO constructs tested in Figure 4C, or ACPL-GFP as a negative control. We attempted to PCR or qPCR amplify multiple nuclear-and apicoplast-encoded genes in purified ChIP and input samples that had been sonicated to shear DNA into fragments ≤2-kb in size prior to IP (Figure 5-figure supplement 4). Target nuclear and apicoplast genes were both successfully amplified in all input samples. However, only the anti-GFP pulldown from parasites expressing full-length PfHO-GFP showed robust amplification of an apicoplast-but not nuclear-encoded gene (Figure 5B and Figure 5 – figure supplements 5 and 6A). Although the portion of the PfHO N-terminus retained in mature PfHO has substantial electropositive character (Figure 1A and Figure 1B) favorable for association with DNA, this sequence in the PfHO1–83-GFP construct was not sufficient for stable pull-down of apicoplast DNA. A faint amplicon for apicoplast DNA was detected for ACPL-PfHO84– 305–GFP, but this signal was >4-fold weaker than observed for full-length PHO (Figure 5B and Figure 5 – figure supplements 5 and 6A). Based on these observations, we conclude that PfHO associates with the apicoplast genome and that DNA-binding requires both the cognate N-terminus and HO-like domain.

HO proteins in other species are reported to bind nuclear DNA53, but the sequence features and nature of those associations are unclear. We note that our ChIP-PCR experiments cannot distinguish whether PfHO pull-down with the apicoplast genome reflects direct association with DNA and/or indirect interactions mediated by other proteins. Nevertheless, the unique sequence features of the PfHO N-terminus and its requirement for DNA association may suggest a Plasmodium-specific mechanism of DNA interaction that differs from other organisms. Selective interaction of full-length PfHO with the apicoplast genome was independent of the target gene amplified by PCR (Figure 5 – figure supplement 5) or qPCR and persisted in the absence of crosslinking (Figure 5 – figure supplement 6). Our observation that 12 distinct genes spanning the apicoplast genome show similar amplification in sheared PfHO ChIP samples (Figure 5 – figure supplements 5 and 6) suggests that PfHO broadly interacts with apicoplast DNA in a sequence-independent manner akin to DNA topology regulators, gyrases, ligases, and single-strand stabilizing proteins70–72, which our IP/MS data suggest are key interactors of PfHO (see discussion below).

Our observation that full-length PfHO, containing the cognate N-terminus and HO-like domain, was concordantly required for both DNA binding (Figure 5B) and essential function (Figure 4C) suggested most simply that association with the apicoplast genome was critical to PfHO function. To test possible roles for PfHO in DNA replication and/or RNA expression, we synchronized parasites to a 5-hour window and determined the impact of PfHO knockdown on DNA and RNA abundance by qPCR and RT-qPCR, respectively, in the first and second cycles after aTC washout but before apicoplast loss and parasite death. Parasites were grown in the presence of 200 µM IPP to decouple cellular viability from apicoplast-specific defects. We observed that PfHO knockdown in -aTC conditions caused a modest ∼20% decrease in apicoplast DNA levels in second-cycle parasites relative to +aTC conditions (Figure 5C). In contrast, apicoplast RNA levels were strongly reduced upon PfHO knockdown, with a 30% reduction observed in the first cycle and nearly 90% reduction in the second cycle after aTC washout (Figure 5D). This PfHO-dependent reduction in RNA abundance was observed for all tested protein-coding and non-coding apicoplast genes spanning all of the currently known or predicted polycistronic apicoplast transcripts (Figure 5 – figure supplement 7)73–76. In contrast to its impact on apicoplast RNA, PfHO knockdown had no measurable impact on RNA transcript abundance for nuclear or mitochondrial genes (Figure 5 – figure supplement 7).

To dissect the time-course of this apicoplast-specific defect in RNA abundance, we collected samples from tightly synchronized parasites throughout the first and second growth cycles after aTC washout. In +aTC conditions, apicoplast RNA levels peaked around 36 hours for most genes, consistent with prior studies of apicoplast transcription77–79. In -aTC conditions, there was a modest decrease in RNA transcript levels in the first cycle but complete failure to increase RNA abundance in the second cycle (Figure 5E and Figure 5 – figure supplement 8). We conclude that PfHO function is essential for apicoplast gene expression and that PfHO knockdown results in a specific defect in RNA abundance that underpins higher-order defects in apicoplast biogenesis that lead to parasite death (Figure 6).

Model for essential PfHO function in apicoplast genome expression and organelle biogenesis. TP = transit peptide. Scissors represent proteolytic processing of the PfHO N-terminal TP upon apicoplast import.

Supplementary file 1. Table of PCR primers.

Supplementary file 2. Table of reagents

Discussion

Heme metabolism is a central cellular feature and critical therapeutic vulnerability of blood-stage malaria parasites, which have evolved unusual molecular adaptations to survive and grow within heme-rich red blood cells. Hemozoin is the dominant fate of labile heme released from large-scale hemoglobin digestion. Nevertheless, malaria parasites express a divergent and inactive heme oxygenase-like protein, whose cellular function underpinning its evolutionary retention has remained mysterious. We have elucidated an essential role for PfHO within the apicoplast organelle of P. falciparum, where it associates with the apicoplast genome and nucleic acid metabolism enzymes and is required for organelle gene expression and apicoplast biogenesis.

Molecular function of PfHO

Our study unveils that P. falciparum parasites have repurposed the HO scaffold from its canonical role in heme degradation towards a divergent function required for expression of the apicoplast genome. This essential function appears to involve direct and/or indirect association with both the apicoplast genome and a variety of DNA/RNA-metabolism enzymes. Also, both the cognate N-terminal leader sequence, most of which remains attached in the mature protein, and the HO-like domain are required for function. We hypothesize that the electropositive N-terminus (Figures 1A and 5B) may mediate direct association with the apicoplast genome, while an electronegative face on the HO-like domain opposite the heme-binding region (Figure 1 – figure supplement 5) may interact with other DNA-binding proteins (e.g., gyrases and helicases) which co-purified with PfHO in IP/MS studies (Figure 6).

The specific molecular function of PfHO that impacts RNA transcript levels in the apicoplast remains to be defined, along with the broader prokaryotic-like biochemical processes that enable transcription in the apicoplast. The apicoplast genome is organized into two polycistronic operons that each consist of roughly half the genome, orient in opposite directions, and contain duplicated rRNA genes at their 5’ ends76. No promotor sequences have been identified in the apicoplast genome, but detection of transcripts up to 15 kb73 and studies on mRNA processing sites74 suggest that full polycistronic operons are transcribed and undergo “punctuation processing” at tRNAs distributed throughout transcripts80,81. The primary subunits of a prokaryotic-like RNA polymerase complex have been identified in Plasmodium. Although associated sigma factors and other interacting proteins are proposed to exist, they have thus far remained difficult to identify73,75. We initially considered a model whereby PfHO might function as a sigma factor-like adaptor that directly binds to the multi-subunit prokaryotic RNA polymerase to mediate genomic association and transcription initiation. However, none of the core RNA polymerase subunits co- purified with PfHO in our IP/MS studies, suggesting that such a function may be unlikely.

Based on PfHO association with a variety of DNA-binding proteins and topology regulators, we consider it more likely that PfHO functions in association with other effector proteins to regulate and optimize DNA topology to enable transcriptional activity. The apicoplast-targeted gyrase A, gyrase B, DDX21 DEAD-box helicase, ligase I, the helicase domain of PREX, histone-like protein (HU)82, and single-strand DNA binding protein71 all co-purify with PfHO in our IP/MS experiments (Figure 5A). Of these proteins, DNA gyrase A83, gyrase B84, PREX85, and HU86 have been directly tested as essential apicoplast biogenesis factors in Plasmodium parasites. The relationship of most of these proteins with apicoplast transcription is unknown, but their activity in other organisms may shed light on a functional pathway in Plasmodium. Prokaryotic transcription is blocked by chemical or genetic disruption of DNA gyrase87–89, a hetero-tetramer of gyrase A and B proteins90. DNA gyrase-induced negative supercoiling locally unwinds circular DNA molecules89 to facilitate transcriptional initiation by RNA polymerase complexes87,91. DNA gyrase also relieves torsional stress formed by the procession of RNA polymerase complexes along DNA molecules and prevents stalling during transcription of long, polycistronic operons92,93. Additionally, HU coordinates prokaryotic DNA structure and supercoiling in conjunction with DNA gyrase and has been reported to regulate the spatial distribution of RNA polymerase and transcription levels in E. coli94,95.

Although these proteins have exclusively been associated with apicoplast DNA replication in Plasmodium, regulation of DNA supercoiling is also a major mechanism of prokaryotic transcriptional control92,96. This biological property may have been present in ancestral plastids and inherited by the prokaryotic-like Plasmodium apicoplast.

We also noted that RNA processing and translation-associated proteins co-purified with PfHO, including ribosomal RNA (rRNA) and transfer RNA (tRNA) maturation proteins, elongation factors, and ribosomal protein subunits (Figure 5A). RNA metabolism in the Plasmodium apicoplast is sparsely understood. No RNA-degrading enzymes have been identified, and the specific functions of RNA-binding proteins remain unknown. DEAD-box DNA/RNA helicases such as DDX21 have been implicated in the removal of aberrant R-loops (DNA/RNA hybrids) during RNA transcription97–99, but Plasmodium DDX21 has also been implicated with rRNA maturation in ribosome biogenesis37. Other RNA-binding proteins that co-purified with PfHO, Nop52 and RAP, also show low-level sequence homology to ribosome assembly factors100,101. Prokaryotic ribosome assembly is a co-transcriptional process regulated by, and in close proximity to RNA transcriptional machinery102–104. Apicoplast ribosome assembly is poorly studied but appears to be similar to prokaryotic systems. Indeed, a prior IP/MS study reported that apicoplast ribosomes and ribosomal assembly complexes co-purified with RNA transcription complexes and DNA topology regulators37. PfHO pulldown with RNA-processing and protein-translation components may therefore reflect the underlying physical and temporal coupling of RNA transcription and protein translation in the apicoplast. It is also possible that PfHO contributes to other aspects of RNA metabolism that remain undefined.

Translation of apicoplast-encoded proteins is required for organelle biogenesis and inheritance105–107, including a likely essential role for the apicoplast-encoded chaperone ClpM in import of nuclear-encoded apicoplast-targeted proteins108–110. Apicoplast-encoded SufB is also expected to be essential for synthesis of Fe-S clusters by the apicoplast SUF pathway that are required for IP synthesis and organelle maintenance111–113. Therefore, we predict that apicoplast biogenesis defects resulting from PfHO knockdown are due to significantly diminished levels of apicoplast-encoded ribosomal and messenger RNAs required for apicoplast translation of ClpM and SufB.

Evolution of PfHO and its divergent function

Although PfHO retains the conserved α-helical structure of HO enzymes, it has strikingly low sequence similarity, implying substantial evolutionary distance from well-studied HO orthologs in humans, plants, and bacteria (Figure 1 and Figure 1 – figure supplement 2). What then is the evolutionary origin of PfHO and its non-canonical function? The Plasmodium apicoplast is thought to derive from secondary endosymbiosis through ancestral engulfment of a plastid-containing red algae that had previously engulfed a photosynthetic cyanobacterium65,114–117. Subsequent loss of photosynthesis accompanied the transition to intracellular parasitism by proto-apicomplexan ancestors109,118. HO enzymes are commonly found in photosynthetic cyanobacteria and eukaryotic chloroplasts, where they initiate biosynthesis of biliverdin and other bilin chromophores utilized in phytochrome proteins for light sensing and signaling14,119,120. Retention of PfHO in the Plasmodium apicoplast likely reflects the original presence of HO enzymes in the cyanobacterial ancestors of this organelle.

The transition from free-living algae to apicomplexan parasitism involved significant genome reduction, including loss of plastid photosynthesis and phytochrome biosynthesis pathways118. These functional reductions presumably removed the selective pressure to retain enzymatic HO activity. Indeed, the only identifiable HO homologs retained in Apicomplexa are found exclusively in hematozoan parasites such as Plasmodium, Babesia, and Theileria, and have lost the active-site His ligand like PfHO (Figure 1 – figure supplement 2). Insights into the evolutionary origin of PfHO are provided by comparison to HO-like proteins in coral-symbiotic chromerid algae, the closest free-living and photosynthetic relatives to apicomplexan parasites109,118. Vitrella brassicaformis and Chromera velia both express multiple HO

homologs thought to be remnants of prior endosymbioses121. These HO proteins segregate phylogenetically with either active metazoan, active plant, or inactive hematozoan HOs (Figure 1 – figure supplement 2) featuring the retention or loss of the active-site His ligand, respectively. Similarly, Arabidopsis thaliana expresses four chloroplast-targeted HO homologs that include three active HO enzymes and the inactive AtHO2 that lacks the conserved His ligand but contributes to photomorphogenesis14. These observations support a model in which active HO homologs were lost along with photosynthetic machinery during the transition to apicomplexan parasitism with the exclusive retention of an inactive HO-like homolog in Hematozoa.

It remains unclear why hematozoan but no other apicomplexan parasites retained an inactive HO- like homolog. The common infection of heme-rich RBCs by hematozoans may suggest that remnant heme- or porphyrin-binding activity8 may play a role in functional regulation of these HO-like proteins within the apicoplast. Indeed, apicoplast transcription78, heme released by hemoglobin digestion122, and PfHO expression (Figure 3- figure supplement 5) all peak coincidentally around 30 hours post-infection. The affinity of PfHO for heme (Kd ∼8 µM)8 is also notably similar to the ∼2 µM concentration of labile heme estimated in the parasite cytoplasm123. Future studies can test the intriguing hypothesis that labile heme levels sensed by PfHO in the apicoplast tune functional interactions by PfHO that regulate apicoplast gene expression.

Although PfHO has diverged from canonical HO function, it retains many structural and biochemical features present in metazoan HOs which are reported to mediate transcription factor activity independent of heme degradation15–17,53,124. Thus, a role for PfHO in gene expression may be an ancient functional property of the HO scaffold that was further expanded and honed by parasites after heme-degrading activity was lost. In this regard, PfHO may be conceptually similar to other parasite proteins (e.g., mitochondrial acyl carrier protein56) that have lost canonical function but whose retention of an essential role unveils a latent secondary activity that was previously un- or under-appreciated in the shadow of the dominant primary function in well-studied orthologs from other organisms.

Implications for expanded functions of N-terminal pre-sequences beyond organelle targeting

Our discovery that PfHO requires a portion of its N-terminal transit peptide for essential function within the apicoplast expands the molecular paradigm for understanding the evolution and function of organelle-targeting leader sequences. PfHO homologs in Vitrella and Chromera also contain a predicted α-helix in their N-terminal sequences that differ from chloroplast-targeting HOs in plants (Figure 4 – figure supplement 4). This unique α-helix predicted in alveolate HO-like proteins adds positive-charge to the HO-like protein that we propose mediates interaction with DNA. It is possible that the alveolate ancestors of Plasmodium expanded and functionally repurposed the targeting sequences of other plastid and/or mitochondrial proteins to provide unique organelle-specific functions that remain to be discovered. The recent report that Toxoplasma gondii parasites repurpose the cleaved leader sequence of mitochondrial cyt c1 as a stable subunit of ATP synthase supports this view125,126. Identifying adaptations in P. falciparum that diverge from human host cells can reveal novel parasite vulnerabilities that underpin the development of new therapeutic strategies.

Materials and methods

Sequence homology searches and phylogenetic analyses

We acquired Plasmodium and alveolate protein sequences from VEuPathDB.org127 and all other protein sequences from UniProt.org128 databases. The protein sequence for PfHO (Pf3D7_1011900) was analyzed by NCBI Protein BLAST19 and HMMER20 with the exclusion of Plasmodium or apicomplexan organisms to identify putative orthologs. Orthologous protein sequences and select reference HO proteins were aligned via Clustal Omega129 and analyzed using Jalview130. The multi-sequence alignment was uploaded to the IQ-TREE webserver (http://iqtree.cibiv.univie.ac.at) with ultrafast bootstrap analysis. The resulting maximum likelihood phylogenetic tree from 1000 bootstrap alignments was analyzed and displayed using FigTree (http://tree.bio.ed.ac.uk/software/figtree/).

Recombinant protein expression and purification for crystal structure determination

The gene encoding residues 84-305 of PfHO was cloned into a pET21d expression vector (Novagen) using NcoI and XhoI sites, in frame with a C-terminal His6 tag8. E. coli BL21 (DE3) cells transformed with this vector were grown in LB medium supplemented with ampicillin (50 µg/mL) (Sigma A9518) and protein expression was induced with 0.5 mM isopropyl β-D-1-thiogalactopyranoside (IPTG) (Sigma 16758) at an OD of 0.5, after which the cells were grown at 20 °C overnight. Cells were harvested by centrifugation (7,000 × g, 10 min), the pellet was resuspended in binding buffer (20 mM Tris-HCl, 500 mM NaCl, 20 mM imidazole, pH 8.5), and the cells were lysed by sonication. The cell lysate was cleared by centrifugation (40,000×g, 30 min), and the supernatant was subjected to immobilized metal affinity chromatography using a 1 mL Ni-NTA agarose column (Qiagen, 30210) equilibrated with binding buffer. The protein was eluted with a buffer containing 20 mM Tris-HCl, 500 mM NaCl, and 300 mM imidazole at pH 8.5. The eluted protein was further purified by size-exclusion chromatography employing a 26/60 Superdex75 column equilibrated with a buffer containing 20 mM HEPES and 300 mM NaCl at pH 7.5. To produce the protein containing SeMet, E. coli B834 (DE3) cells transformed with the same plasmid as above were grown in minimal media (2 g L-1 NH4Cl, 6 g L-1 KH2PO4, 17g L-1 Na2HPO4·12H2O, 1 g L-1 NaCl, 1.6 mg L-1 FeCl3, 0.5 g L-1 MgSO4·7H2O, 22 mg L-1 CaCl2·6H2O, 4 g L-1 Glucose, and 50 mg L-1 L-SeMet). Purification was identical to that of the WT protein. Protein purity was confirmed by observation of a single band at the appropriate molecular mass by Coomassie-stained SDS-PAGE.

Protein crystallization

Protein in a buffer containing 20 mM HEPES and 300 mM NaCl at pH 7.5 was subjected to crystallization trials using sitting-drop vapour diffusion using commercially available screening kits form Hampton Research in an Oryx8 system (Douglas Instruments). Protein (0.5 μL at 9.5 mg mL-1) was mixed with an equal volume of reservoir solution, and crystallization plates were maintained at 20 °C for several weeks while being examined. The crystallization solutions producing the best crystals were optimized using hanging-drop geometry in 24-well plates by mixing manually 2 µL of protein solution (5.0 mg mL-1) and an equal volume of reservoir solution. The best crystals appeared in a few days in a reservoir solution containing 0.4 M (NH4)2SO4, 0.65 M Li2SO4, and 0.1 M sodium citrate tribasic dihydrate at pH 5.6 and a temperature of 20 °C. Single crystals were mounted in nylon Cryo-Loops (Hampton Research, HR4-932) coated with Paratone (Hampton Research, HR2-862) and directly transferred to liquid nitrogen for storage.

Structural data collection and processing

Diffraction data from single crystals of WT and SeMet protein were collected in beamlines AR-NW12A and BL5A, respectively, at the Photon Factory (Tsukuba, Japan) under cryogenic conditions (100 K). Diffraction images were processed with the program MOSFLM and merged and scaled with the program SCALA or AIMLESS of the CCP4 suite. The structure of the SeMet protein was solved by the method of single anomalous diffraction using the Autosol module included in the PHENIX suite. The structure of the WT protein was determined by the molecular replacement method using the coordinates of the SeMet protein from above with the program PHASER. The models were refined with the programs REFMAC5 and built manually with COOT. Validation was carried out with PROCHECK. Data collection and structure refinement statistics are given in Figure 1 – figure supplement 3. The final structural coordinates and structure factors were deposited as RCSB Protein Data Bank entry 8ZLD.

Structural visualization and analyses

A predicted structural model for PfHO (Pf3D7_1011900) was acquired from the AlphaFold Protein Structure Database (https://alphafold.ebi.ac.uk), and published HO structures were acquired from the RCSB Protein Data Bank (https://www.rcsb.org). The AlphaFold structural model, PfHO crystal structure, and HO structures were visualized and analyzed using The PyMOL Molecular Graphics System, Version 2.5, Schrödinger, LLC. Structural superpositions were performed with the PyMOL integrated command “align” and assessed by the total number of atoms aligned and RMSD (Å). We uploaded PDB files of PfHO crystal structure and AlphaFold model to the DALI protein structural comparison server (http://ekhidna2.biocenter.helsinki.fi/dali) to identify proteins structurally related to PfHO. To determine surface charge of structures, we uploaded PDB files to APBS-PDB2PQR online software suite (https://server.poissonboltzmann.org/)131, and displayed the calculated Poisson-Boltzmann surface charge using the PyMOL APBS tool 2.1 plugin.

Parasite culturing and transfection

Plasmodium falciparum Dd2132 or 3D742 parasites were cultured in human erythrocytes obtained from Barnes-Jewish Hospital (St. Louis, MO) or the University of Utah Hospital blood bank (Salt Lake City, UT) in RPMI-1640 medium (Thermo Fisher, 23400021) supplemented with 2.5 g/L Albumax I Lipid-Rich BSA (Thermo Fisher, 11020039), 15 mg/L hypoxanthine (Sigma, H9636), 110 mg/L sodium pyruvate (Sigma, P5280), 1.19 g/L HEPES (Sigma, H4034), 2.52 g/L sodium bicarbonate (Sigma, S5761), 2 g/L glucose (Sigma, G7021), and 10 mg/L gentamicin (Invitrogen, 15750060), as previously described49. Parasites were maintained at 37° C in 90% N2/5% CO2/5% O2 or in 5% CO2/95% air. For drug-induced apicoplast-disruption experiments, parasites were cultured for ∼7 days in 1 µM doxycycline (Sigma, D9891) and 200 µM isopentenyl pyrophosphate (Isoprenoids, IPP001).

Transfections were performed in 1x cytomix containing 50-100 µg DNA by electroporation of parasite-infected RBCs in 0.2 cm cuvetes using a Bio-Rad Gene Pulser Xcell system (0.31 kV, 925 µF). Transgenic parasites were selected based on plasmid resistance cassettes encoding human DHFR133, yeast DHOD134, or blasticidin-S deaminase135 and cultured in 5 nM WR99210 (Jacobus Pharmaceutical Co.), 2 µM DSM1 (Sigma, 53330401), or 6 µM blasticidin-S (Invitrogen, R21001), respectively. After stable transfection and selection, parasites were grown in the continual presence of selection drugs, and aptamer-tagged parasites were grown in 0.5-1 µM anhydrotetrocycline (Cayman Chemicals, 10009542).

Parasite growth assays

Parasites were synchronized to the ring stage with an estimated 10-15-hour synchrony window by treatment with 5% D-sorbitol (Sigma, S7900). For aptamer-based knockdown experiments, aTC was washed out during synchronization with additional 3-5 washes in media and/or PBS. Parasite growth was monitored by plating synchronized parasites at ∼1% parasitemia and allowing culture expansion over several days with daily media changes. Parasitemia was monitored daily by flow cytometry by diluting 10 µL of each parasite culture well from each of three biological replicate samples into 200 µL of 1.0 µg/mL acridine orange (Invitrogen, A3568) in phosphate buffered saline (PBS) then analyzed on a BD FACSCelesta flow cytometry system monitoring SSC-A, FSC-A, PE-A, FITC-A, and PerCP-Cy5-5-A channels.

Cloning and episomal expression of PfHO variants in parasites

The genes encoding PfHO (Pf3D7_1011900) and apicoplast ACP (Pf3D7_0208500) were PCR amplified from P. falciparum strain 3D7 cDNA using primers with ≥20 bp homology overhangs (primers 10-15) for ligation-independent insertion into the XhoI and AvrII sites of pTEOE (human DHFR selection cassette) in frame with a C-terminal GFP tag, and with expression driven by HSP86 promoter136. Correct plasmid sequences were confirmed by Sanger sequencing, and plasmids were transfected as described above in combination with 25 µg pHTH plasmid containing piggyBac transposase to drive stable, random integration into the parasite genome137.

Parasite genome editing to enable ligand-dependent regulation of PfHO expression

We first used restriction endonuclease-mediated integration39 and single-crossover homologous recombination to tag the PfHO gene to encode a C-terminal GFP-tag fused to the DHFR-destabilization domain40,41 and a single hemagglutinin (HA) tag in 3D7 (PM1 KO)42 parasites. PCR primers 18 and 19 were used to clone the 3’ 1kb DNA sequence of the PfHO gene into the XhoI and AvrII sites of the pGDB vector40 to serve as a homology region for integration. 50 µg of this plasmid along with 50 units of MfeI restriction enzyme (NEB R3589), which cuts at a single site within the PfHO gene just upstream of the homologous sequence cloned into the donor-repair pGDB plasmid, was transfected into 3D7 PM1 KO parasites42 (which express human DHFR), as described above. Parasites were positively selected with blasticidin-S in the continuous presence of trimethoprim (Sigma, T7883), cloned by limiting dilution, and genotyped by probing Southern blots of MfeI/HindIII-digested total parasite DNA with a gel-purified, 1 kb PCR product of the 3’ UTR of PfHO (Figure 3-figure supplement 1). Southern blot signal was generated with an AlkPhos direct labeling and detection kit as previously described49,138. Conditional knockdown was evaluated by synchronized parasite growth assays after 3-5x washes in PBS to remove trimethoprim.

We next used CRISPR/Cas9 and single-crossover homologous recombination to tag the PfHO gene to encode a C-terminal HA2 tag fused to the glmS ribozyme43,44 in Dd2 parasites. The 1 kb homology sequence or PfHO was excised from pGDB and sub-cloned by ligation into a modified pPM2GT vector138 in which the linker-GFP sequence between the AvrII and EagI sites was replaced with a HA-HA tag and stop codon followed by the 166 bp glmS ribozyme43,44. QuikChange II site-directed mutagenesis (Agilent Technologies) was used with primers 20 and 21 to introduce silent shield mutations into the PfHO homology region for purposes of CRISPR/Cas9-based genome editing, such that the AGATGG sequence in the most 3’ exon was changed to CGGTGG. A guide RNA sequence corresponding to TGAGTAGGAAATGGAGTAGA was cloned into a modified version of the previously published pAIO CRISPR/Cas9 vector139 in which the BtgZI site was replaced with a unique HindIII site to facilitate cloning49. 50 µg each of the pPM2GT and pAIO vectors were transfected into Dd2 parasites, as described above. Parasites were positively selected by WR99210, cloned by limiting dilution, and genotyped by PCR (Figure 3 – figure supplement 1). Conditional knockdown was evaluated by adding 0-10 mM glucosamine (Sigma, G1514) to synchronized parasite cultures and evaluating protein expression and parasite growth relative to untagged parental Dd2 parasites.

CRISPR/Cas9 and double-crossover homologous recombination was used to tag the PfHO gene to encode a single RNA aptamer at the 5’ end and a 10x aptamer cassette at the 3’ end for inducible knockdown with the aptamer/TetR-DOZI system46,47. A donor plasmid was created by ligation-independent insertion of a synthetic gene (gBlock, IDT) containing PfHO cDNA (T. gondii codon bias) into the linear pSN1847L vector46, along with PCR amplified (primers 22-25) left and right homology flanks corresponding to the 5’ (426 bp immediately upstream of start codon) and 3’ (455 bp starting at position 47 after the TAA stop codon) untranslated regions of PfHO. Because the aptamer sequence contains two ATG motifs that can serve as alternate translation start sites, a viral 2A peptide sequence was introduced between the 5’ aptamer sequence and the start ATG of PfHO. This donor repair plasmid (50 µg) and the pAIO CRISPR/Cas9 vector (50 µg) with guide sequence TGAGTAGGAAATGGAGTAGA targeting the 3’ end of the endogenous PfHO gene was transfected into Dd2 parasites, as described above. No shield mutation in the donor plasmid was required due to the altered codon bias of the synthetic PfHO cDNA in that vector. Parasites were positively selected with blasticidin-S in the presence of 1 µM aTC. Integration was confirmed by PCR and probing Southern blots of NdeI/HindIII-digested total parasite DNA with a gel-purified, 580 bp PCR product of the 3’ UTR of PfHO (Figure 3 – figure supplement 4). Southern blot signal was generated with an AlkPhos direct labeling and detection kit as previously described49,138. Parasites were also cloned by limiting dilution, however no evidence of remnant WT locus was detected in polyclonal transfectants and identical phenotypes were observed for polyclonal and clonal parasite cultures. Therefore, polyclonal cultures were used for all subsequent experiments.

Microscopy

Live microscopy of parasites expressing GFP-tagged proteins was performed by staining mitochondria with 25 nM MitotrackerRed CMXRos (Invitrogen, M7512) for 30 minutes and staining nuclei with 5 µg/mL Hoechst 33342 (Invitrogen, H3570) for 5-10 minutes in PBS. Stained parasites were then imaged in PBS under a coverslip on an Invitrogen EVOS M5000. Images were adjusted for brightness and contrast in FIJI with linear variations equally applied across images. Signal intensity profiles were calculated for the red and green channels respectively using the FIJI plot profile tool along a single line that transects the region of highest signal for both channels (identified on images as white line). At least 20 individual parasites were imaged for each parasite line or each condition.

For immunofluorescence (IFA) experiments, parasitized red blood cells were fixed in 4% paraformaldehyde and 0.0016% glutaraldehyde for 30 minutes at 25°C, then deposited onto poly-D-lysine coated coverslips. Fixed cells were permeabilized in PBS supplemented with 0.1% Triton-X100, reduced in 0.1 mg/mL NaBH4, and blocked in 3% BSA for 30 minutes. Parasites were stained with primary antibodies: mouse anti-GFP (Invitrogen, A-11120), and rabbit anti-apicoplast ACP140 at 1:100 dilution in 1% BSA for 1 hour at 25°C, washed thrice in PBS-T (PBS with 0.1% Tween-20), stained with secondary antibodies: goat anti-mouse AF488 (Invitrogen, A-11001) and goat anti-rabbit AF647 (Invitrogen, A21244) in 1% BSA for 1 hour at 25°C, and washed thrice in PBS-T before imaging. Coverslips were mounted onto slides using ProLong Diamond Antifade Mountant with DAPI (Invitrogen, P36971) overnight at 25°C, then imaged on an Axio Imager M1 epifluorescence microscope (Carl Zeiss Microimaging Inc.) equipped with a Hamamatsu ORCA-ER digital CCD camera. Images were adjusted for brightness and contrast in FIJI with linear variations equally applied across images. Pearson correlation was calculated with the FIJI Coloc2 tool on unmasked images using a point spread function of 3 pixels and 50 Costes iterations. At least 10 individual parasites were imaged for each parasite line.

Immunogold transmission electron microscopy was performed (Dr. Wandy Beatty, Washington University in St. Louis) as previously described141 using endogenously tagged PfHO-GFP-DHFRDD 3D7 parasites and staining with goat anti-GFP (Abcam, ab5450) and rabbit anti-apicoplast ACP142 antibodies along with gold-conjugated anti-goat and anti-rabbit secondary antibodies. 15 individual parasites were imaged.

Production and validation of custom anti-PfHO rabbit antibody

The HO domain of PfHO (84-305) was cloned into pET28 with an N-terminal His-tag and start codon, purified by Ni-NTA, cleaved by thrombin, and purified by FPLC as previously reported8. Purified protein was then injected into a rabbit for polyclonal Ab production by Cocalico Biologicals Inc. (https://www.cocalicobiologicals.com) following their standard protocol. Rabbit serum was validated for specific detection of PfHO in parasite lysates and with recombinant protein expressed in bacteria prior to use (Figure 3 – figure supplement 5).

SDS-PAGE and western blots

Parasite cultures were grown to ∼10% parasitemia in 10 mL cultures for western blots and 50-100 mL cultures for immunoprecipitation. Parasites were released from red blood cells by treatment with 0.05% saponin and subsequently lysed by sonication (20 pulses 50% duty cycle, 50% power) on a Branson microtip sonicator in TBS lysis buffer (50mM Tris pH 7.4, 150 µM NaCl, 1% v/v Triton X-100) with protease inhibitors (Invitrogen, A32955) followed by incubation at 4°C for 1 hour. Recombinantly expressed protein was obtained for western blot analysis by inducing BL21 (DE3) E. coli grown to an OD of 0.5 in LB medium with 0.5 mM IPTG at 20 °C overnight. Cells were harvested by centrifugation (5,000 RPM, 10 min), and lysed with the same methods described above. Lysates were clarified by centrifugation (14,000 RPM, 10 min) and quantified by Lowry colorimetry. 50 µg of total protein was mixed in SDS sample buffer, heated at 95C for 10 minutes, and separated by electrophoresis on 12% SDS-PAGE gels in Tris-HCl buffer. Proteins were transferred onto nitrocellulose membranes using the Bio-Rad wet-transfer system for 1 hour at 100 V, and blocked with 5% non-fat milk in TBS-T (50mM Tris pH 7.4, 150 µM NaCl, 0.5% v/v Tween-20). Membranes were stained with primary antibodies: goat anti-GFP (Abcam ab5450), rat anti-HA (Roche), mouse anti-hDHFR, rabbit anti-apicoplast ACP140, custom rabbit anti-PfHO, or rabbit anti-EF1α143 diluted 1:1000-1:2500 in blocking buffer for ≥ 18 hours at 4 °C. Samples were then washed thrice in TBS-T and stained with secondary antibodies: rabbit anti-Goat HRP (Santa Cruz, sc2768), goat anti-Rabbit HRP (Invitrogen, A27036), donkey anti-rabbit IRDye800CW (LiCor, 926-32213), donkey anti-rabbit IRDye680 (LiCor, 926-68023), or donkey anti-goat IRDye800CW (LiCor, 926-32214), diluted 1:10,000 in TBS-T for 1-2 hours at 25 °C, and again washed thrice before imaging. All blots stained with horseradish peroxidase (HRP) were developed 3 minutes with Prometheus™ ProSignal Femto ECL reagent (Genesee Scientific 20-302) and imaged on a BioRad ChemiDoc MP imaging system, and all blots stained with IRDye antibodies were imaged on a Licor Odyssey CLx system. Protein size was estimated by migration relative to the protein ladder using Licor Image Studio software v5.5.4.

N-terminal protein sequencing of PfHO

A 170 mL culture of 3D7 parasites at 3% hematocrit and ∼15% asynchronous parasitemia was harvested by centrifugation followed by release of parasites from RBCs in 5% saponin (Sigma S7900). The resulting parasite pellet was subsequently lysed in 1 mL RIPA buffer with sonication and clarified by centrifugation. PfHO was isolated by immunoprecipitation using 60 µl of affinity-purified custom anti-PfHO rabbit antibody and 400 µl of Protein A dynabeads (Invitrogen, 1001D). The beads were washed 3X in RIPA buffer, eluted with 110 µl 1X SDS sample buffer, fractionated by 10% SDS-PAGE, followed by transfer to PVDF membrane and staining by Coomassie. The band corresponding to mature PfHO was excised and subjected to N-terminal sequencing by Edman degradation at the Stanford University Protein and Nucleic Acid Facility.

Immunoprecipitation

Dd2 parasites expressing endogenously tagged PfHO with C-terminal HA-HA tag were harvested from ∼75 mL of culture by centrifugation, released from RBCs by incubating in 0.05% saponin (Sigma 84510) in PBS for 5 minutes at room temperature, and pelleted by centrifugation (5000 rpm, 30 min, 4°C). Parasites were then lysed by sonication (20 pulses 50% duty cycle, 50% power) on a Branson microtip sonicator in TBS lysis buffer (50mM Tris pH 7.4, 150 µM NaCl, 1% v/v Triton X-100) with protease inhibitors (Invitrogen A32955) followed by incubation at 4°C for 1 hour and centrifugation (14,000 rpm, 10 min), The clarified lysates were mixed with equilibrated resin from 30 µl of Pierce anti-HA-tag magnetic beads (Invitrogen 88836) and incubated for 1 hour at 4°C on a rotator. Beads were placed on a magnetic stand, supernatants were removed by aspiration, and beads were washed thrice with cold TBS-T. Bound proteins were eluted with 100 µl of 8M urea (in 100 mM Tris-HCl at pH 8.8). Proteins were precipitated by adding 100% trichloroacetic acid (Sigma 76039) to a final concentration of 20% v/v and incubated on ice for 1 hour. Proteins were then pelleted by centrifugation (13,000 rpm, 25 min, 4°C) and washed once with 500 µl of cold acetone. The protein pellets were air-dried for 30 min and stored at −20°C.

Mass spectroscopy

Protein samples isolated by anti-HA-tag IP of endogenous PfHO were reduced and alkylated using 5 mM Tris (2-carboxyethyl) phosphine and 10 mM iodoacetamide, respectively, and then enzymatically digested by sequential addition of trypsin and lys-C proteases, as previously described144. The digested peptides were desalted using Pierce C18 tips (Thermo Fisher Scientific), dried, and resuspended in 5% formic acid. Approximately 1 μg of digested peptides was loaded onto a 25-cm-long, 75-μm inner diameter fused silica capillary packed in-house with bulk ReproSil-Pur 120 C18-AQ particles, as described previously145. The 140-min water-acetonitrile gradient was delivered using a Dionex Ultimate 3,000 ultra-high performance liquid chromatography system (Thermo Fisher Scientific) at a flow rate of 200 nl/min (Buffer A: water with 3% DMSO and 0.1% formic acid, and Buffer B: acetonitrile with 3% DMSO and 0.1% formic acid). Eluted peptides were ionized by the application of distal 2.2 kV and introduced into the Orbitrap Fusion Lumos mass spectrometer (Thermo Fisher Scientific) and analyzed by tandem mass spectrometry. Data were acquired using a Data-Dependent Acquisition method consisting of a full MS1 scan (resolution = 120,000) followed by sequential MS2 scans (resolution = 15,000) for the remainder of the 3-s cycle time. Data was analyzed using the Integrated Proteomics Pipeline 2 (Integrated Proteomics Applications, San Diego, CA). Data were searched against the protein database from P. falciparum 3D7 downloaded from UniprotKB (10,826 entries) on October 2013. Tandem mass spectrometry spectra were searched using the ProLuCID algorithm followed by filtering of peptide-to-spectrum matches by DTASelect using a decoy database-estimated false discovery rate of <1%. The proteomics data are deposited in the MassIVE data repository (https://massive.ucsd.edu) under the identifier MSV000094692.

Measuring apicoplast DNA and RNA abundance in parasites

Highly synchronous parasites were obtained by sorbitol synchronization of high parasitemia cultures followed by magnet-purification of schizonts after 36-40 hours using MACS LD separation columns (Miltenyi, 130-042-901) with stringent washing. Purified schizonts were allowed to reinvade fresh RBCs for 5 hours on an orbital shaker at 100 rpm in media containing 1 µM aTC. Immediately prior to experimental plating, parasites were treated with sorbitol to ensure ≤5-hour synchrony window and washed 3-5 times in media and/or PBS to remove aTC. Times listed in growth assays are post-synchronization and reflect T=0 at the time that magnet-purified schizonts were allowed to reinvade fresh RBCs.

Highly synchronous parasites were plated in 4 mL of either +aTC or -aTC media. Parasites to be harvested in the first life cycle were cultured at 3%, second cycle at 1%, and third cycle at 0.5% starting parasitemia. 4 mL of +aTC and -aTC cultures were collected for whole DNA and RNA extraction, respectively, at 36, 84, and 132 hours, snap frozen in liquid N2, and stored at -80°C. Parasite DNA was extracted with a QIAamp DNA Blood Mini kit (Qiagen, 51106), and RNA was purified by Trizol (Invitrogen, 15596026) and phenol-chloroform isolation. We converted 1 µg of purified RNA to cDNA using a SuperScript IV VILO RT kit (Invitrogen, 11766050). Since apicoplast genes are extremely AT-rich76 and mRNA transcripts are not poly-adenylated or poly-uridylylated80, gene-specific reverse primers were used to prime the reverse transcription reactions. RT-qPCR was then used to assess the DNA/RNA abundance of four nuclear genes: STL (Pf3D7_0717700), I5P (Pf3D7_0802500), ADSL (Pf3D7_0206700), and PfHO (Pf3D7_1011900), one mitochondrial gene: CytB (Pf3D7_MIT02300), and twelve apicoplast genes: rpl-4 (Pf3D7_API01300), rpl-2 (Pf3D7_API01500), rpl-14 (Pf3D7_API02000), rps-12 (Pf3D7_API02700), EF-Tu (Pf3D7_API02900), ClpM (Pf3D7_API03600), RpoC2 (Pf3D7_API04200), RpoC1 (Pf3D7_API04300), RpoB (Pf3D7_API04400), SufB (Pf3D7_API04700), ls-rRNA (Pf3D7_API06700), and ss-rRNA (Pf3D7_API05700) (primers 28-61). Invitrogen Quantstudio Real-Time PCR systems were used to quantify abundance of DNA and cDNA using SYBR green dye and primers 28-61. The relative DNA or cDNA abundance of each apicoplast gene was normalized to the average of three nuclear-encoded genes for each sample, and -aTC was compared to +aTC by the comparative Ct method146. All qPCR experiments were performed in triplicate and data was analyzed by unpaired Student’s t-test.

PfHO chromatin immunoprecipitation (ChIP) analysis

We saponin-released 75mL of high parasitemia Dd2 cultures transfected with episomes encoding expression of PfHO-GFP, PfHO1–83-GFP, ACPL-PfHO84–305–GFP, and ACPL-GFP and crosslinked in 1% paraformaldehyde for 15 minutes at 20°C, then quenched with 125mM glycine. Crosslinked parasites were transferred into 2 mL ChIP lysis/sonication buffer (200 mM NaCl, 25 mM Tris pH 7.5, 5 mM EDTA pH 8, 1% v/v Triton X-100, 0.1% SDS w/v, 0.5% sodium deoxycholate w/v, and protease inhibitors), and sonicated for 15 cycles of 30s ON/OFF at 25% power using a microtip on a Branson sonicator, then clarified by centrifugation (14,000 rpm, 10 min, 4°C). DNA fragment size after shearing was determined by Agilent Bioanalyzer DNA analysis (University of Utah DNA Sequencing Core) (Figure 5 – figure supplement 4). We collected 200µL (10%) of the clarified, sheared lysates as “input controls” and the incubated the rest with goat anti-GFP antibody (Abcam, ab5450) at 4°C overnight. Antibody-bound protein was mixed with equilibrated resin from 25 µL protein A-conjugated dynabeads (Invitrogen, 1011D) for 1 hour at 4°C rotating, then washed in lysis/sonication buffer, 1 mg/mL salmon sperm DNA (Invitrogen, AM9680) in lysis/sonication buffer, high salt wash buffer (500 mM NaCl, 25 mM Tris pH 7.5, 2 mM EDTA pH 8, 1% v/v Triton X-100, 0.1% w/v SDS, and protease inhibitors), and Tris-EDTA buffer. Samples were eluted from protein A dynabeads by two rounds of 5-minute incubation at 65°C in 100 µL elution buffer (10 mM Tris pH 8, 1 mM EDTA, 1% w/v SDS). We increased NaCl concentration in both input control and ChIP elution samples to 200 mM and added 50 µg/mL RNAse A, then incubated overnight (at least 8 hours) at 65°C to reverse crosslinks and digest RNA. We increased EDTA concentration to 5 mM and added 2 uL of 20 mg/mL Proteinase K and digested at 60°C for 1 hour, then purified DNA using the Qiagen PCR purification kit. Purified DNA was immediately used for either steady-state PCR using primers 62-73 or qPCR amplification using protocol described above. Relative quantification of steady-state PCR bands was performed by area-under-the-curve densitometry analysis in FIJI. In qPCR experiments, amplification of each gene in ChIP DNA was normalized to amplification of the same gene in DNA purified from the input control to account for variability between parasite lines. All densitometry and qPCR experiments were performed in triplicate and statistical significance of differences between PfHO-GFP and other constructs was calculated using Student’s t-test.

Supporting information